Reduced dorsal CA1 Activity Limits Retention of the Temporal Component of Declarative Memory in the Cntnap2 Knockout Mouse Model of Autism

Version 1

Abstract

Growing evidence implicates the hippocampus in the pathophysiology of autism spectrum disorder, particularly in the domains of social interactions and cognition. Yet, the mechanisms driving hippocampal-dependent cognitive atypicalities in autism remain poorly defined. Here, we characterized how dysfunction of the CA1 subfield of the dorsal hippocampus drives critical components of declarative memory. Using trace fear conditioning in the Cntnap2 knockout mouse model of autism, we found that capabilities to retain the association of temporally distant stimuli (i.e. temporal binding) were reduced relative to wildtype mice. Fiber photometry and optogenetic experiments demonstrated that reduced CA1 activity during temporal gaps underlies this impairment, linking CA1 dysfunction to a deficit of long-term memory retention in autism. Using a relational/declarative memory task, we also revealed a deficit in flexible spatial memory, and a preferential use of egocentric learning strategy. This unflexible learning strategy resulted from the imbalance between memory system activity, promoting frontostriatal-dependent procedural learning instead of dorsal CA1/hippocampus-dependent relational/declarative memory. Overall, this study establishes dorsal CA1 dysfunction as a circuit-level mechanism underlying cognitive inflexibility in autism, providing a neurobiological framework for hippocampal-dependent memory deficits in the condition.

Introduction

Autism spectrum disorder (ASD) is a lifelong neurodevelopmental disorder comprising difficulties in social behavior, atypical communication skills, and the presence of restricted or repetitive behaviors or interests (DSM-5 and Association American Psychiatric, 2013). Beyond these core features, individuals with ASD present with difficulties in cognitive domains which can impact daily functioning and adaptability. In particular, people with ASD present with atypicalities in episodic/declarative memory (Bal et al., 2015), the ability to recall specific events and experiences (i.e., what happened, when, and where). Yet, the underlying neurological mechanisms of declarative memory impairments in ASD remain to be fully elucidated.

While most research has focused on the dysregulation of the prefrontal cortex and the basal ganglia in ASD, recent studies have highlighted that hippocampal defects are key in the social domain of this phenotype (Banker et al., 2021). In addition, the hippocampus is known to play a critical role in declarative memory by forming temporally and spatially organized relational representations of past experiences, allowing flexible expression of memories (Eichenbaum, 2004). The CA1 subfield plays a key role in this relational process, by sustaining the ability to link temporally distant events, called temporal binding (Eichenbaum, 2014; Sellami et al., 2017). Difficulties in the binding of episodic memories have been proposed in autistic individuals, positing a selective weakness in hippocampus-dependent linking of items and contexts, while memory for items alone remains relatively spared (Bowler et al., 2011; Minor et al., 2023). In parallel with this relational hypothesis of memory defects, time estimation tasks in individuals with ASD have shown that disruptions in integrating temporal information affect the development of temporal relationships (Buonomano et al., 2023; Maister and Plaisted-Grant, 2011; Vogel et al., 2022). In this context, we hypothesized that CA1 activity underlying temporal processing and temporal linking is impaired in ASD, leading to declarative memory defects.

Here, we examined this hypothesis in the Cntnap2 knockout (KO) mouse model of ASD, which exhibits autism-like phenotypes, including social behavior deficits (Peñagarikano et al., 2011; Peñagarikano and Geschwind, 2012) and hippocampal alterations, including CA1 circuit-level changes and impairments in spatial discrimination (Paterno et al., 2021). Animals were assessed in two behavioral paradigms: trace fear conditioning, a simple temporal binding-dependent memory task allowing the simultaneous recording of CA1 activity; and a radial maze task modeling the complexity of temporal binding-dependent relational/declarative memory. We demonstrate that Cntnap2 KO mice present mistuned dCA1 activation and abnormal reorganization of memory systems during temporal binding-dependent memory formation.

Results

Cntnap2 KO mice demonstrate reduced temporal binding capability

We tested the temporal binding capabilities of the Cntnap2 KO mouse model using the trace fear conditioning paradigm (TFC; Figure 1A), in which mice are required to memorize the association between a tone (Conditioned Stimulus) and a mild electric foot-shock (Unconditioned Stimulus), separated by a time interval called Trace (Sellami et al., 2017). In the first stage, i.e., the conditioning, independent groups of mice were exposed to three tone-shock pairings with a trace interval between 5 and 40 s (i.e. T5, T20, or T40; Figure 1A). All groups showed intact conditioning abilities, demonstrated by high and increasing freezing times (Blanchard et al., 1986) during the tones and traces periods (Figure 1B, Figure S1A). Elevated freezing during the pre-tone period may reflect fear generalization (Figure S1B, T20 vs. T40 (Sellami et al., 2017)). 24h later, the animals were re-exposed first to the conditioned tone in a neutral context (i.e. tone test) to assess 24h retention capacities of the tone-shock association. Retention was reflected in fear responses to the tone and trace compared with before and after the tone (i.e., conditioned fear response in Figure 1B, % time freezing in Figure S1B). The animals were then exposed to the conditioning context alone to evaluate contextual memory. As previously demonstrated (Al Abed et al., 2020b; Sellami et al., 2017), WT mice displayed efficient retention of the tone-shock association up to a 40-s trace interval (Figure 1C, Figure S1B), illustrated by strong and specific conditioned fear response to the tone in a neutral context (Figure 1C, Figure S1B). Interestingly, Cntnap2 KO mice also displayed strong and specific freezing to the tone for up to a 20 s trace interval (Figure 1C, T20), but not for 40 s (Figure 1C, T40; Figure S1B, T40). These results were replicated in both control and KO littermates (Figure S1C). Altogether, these results indicate that Cntnap2 KO mice exhibit reduced temporal binding capacity compared to WT mice, which compromises the 24h retention of the tone-shock association.

Reduced retention of temporal associations is associated with altered calcium responses during trace fear conditioning acquisition in Cntnap2 KO mice

A. Experimental design: Trace Fear Conditioning includes a conditioning session in which mice have to associate a tone and a shock, separated by a time interval (trace). Independent groups of mice were trained with trace intervals of 5, 20, or 40 seconds. B. Freezing levels (percent) during the third tone of the conditioning session for WT (black) and Cntnap2 KO (purple) groups, conditioned with distinct trace durations (T5: WT n=10; Cntnap2 KO n=11; T20: WT n=12; Cntnap2 KO n=11; T40: WT n=12; Cntnap2 KO n=14). C. Conditioned fear response during the tone test (i.e., ratio of freezing before the tone compared to during the tone and the trace, see methods), 24h post-conditioning with distinct trace durations (Trace X Genotype interaction: F2, 65=11.278; p<0.0001; T40: Genotype: F2, 24=6.944; p=0.0163). D.Freezing levels in the conditioning context (context test) for WT (black) and Cntnap2 KO (purple) groups, 24h post-conditioning. E. Fiber photometry. Left panel: scheme depicting the fiber optic implantation targeting the dorsal CA1 subfield of the hippocampus (dCA1). Right panel: Representative image of the optic fiber placement, in the pyramidal cell layer of the dCA1, scale: 100µm. Calcium activity was recorded using the calcium indicator GCaMP6f. F. Freezing levels (percent) during the third tone of the conditioning session for WT (black) and Cntnap2 KO (purple) groups, conditioned with distinct trace durations (T20: WT n=8; Cntnap2 KO n=8; T40: WT n=11; Cntnap2 KO n=10). G. Conditioned fear response during the tone test (i.e., ratio of freezing before the tone compared to during the tone and the trace, see methods), 24h post-conditioning with distinct trace durations (T40: Genotype: F1, 19=50.143; p<0.0001). H. Left: Peak amplitude of Ca2+ transients during the 3 tone presentations of the conditioning under a 20s trace-interval (ns). Right: Peak amplitude of Ca2+ transients during the 3 trace intervals of the conditioning under a 20s trace-interval (Genotype x Trace interaction: p=0.0476; WT n=8; KO n=8, trace 1: ns (p=0675)). I. Left: Peak amplitude of Ca2+ transients during the 3 tone presentations of the T40 conditioning session (Genotype x Tone interaction: p=0.0247; Tone: p=0.0053; WT n=11; KO n=10). Right: Peak amplitude of Ca2+ transients during the 3 trace intervals of the T40 conditioning session (Genotype: p=0.0408; WT n=11; KO n=10). J. Optogenetics: Left panel: scheme depicting the fiber optic implantations bilaterally targeting the dorsal CA1 subfield of the hippocampus (dCA1). Right panel: Representative image of the optic fiber placement, in the pyramidal cell layer of the dCA1, scale: 100µm. Pyramidal neurons activity was manipulated only during the trace interval using Channel Rhodopsin (ChR2). K. Freezing levels (percent) during the third tone of the conditioning session for GFP-injected WT (black, n=6), GFP-injected Cntnap2 KO (purple, n=7), and ChR2-injected Cntnap 2 KO (light purple/white, n=11) groups, conditioned with 40s trace duration. G. Conditioned fear response during the tone test (i.e., ratio of freezing before the tone compared to during the tone and the trace, see methods), 24h post-conditioning with distinct trace durations (Groups: F2, 21=15.576; p<0.0001; WT-GFP vs KO-GFP: p=0.0010; KO-GFP vs KO-ChR2: p<0.0001). Data shown at mean ± SEM. *: p<0.05; **: p<0.01; ***: p<0.001; * represents genotype effect. # represents genotype x variable interaction.

Because freezing is the absence of movement, the well-reported hyperlocomotion in Cntnap2 KO mice ((Peñagarikano et al., 2011); Figure S1D) could represent a potential confounding factor. However, we observed comparable levels of resting time (Figure S1D, right panel) and similar freezing levels between WT and Cntnap2 KO mice during acquisition, regardless of the trace interval (Figure S1A), as well as during memory tests under control conditions (Figure S1B; T5, T20, and T60). These findings indicate that locomotor activity alone cannot account for the observed differences in freezing performance. Taken together, our results demonstrate that Cntnap2 KO mice exhibit reduced temporal binding capacity during conditioning, which does not affect acquisition but impairs the 24h retention of the tone-shock association.

Mistuning of dorsal CA1 activity during temporal binding in Cntnap2 KO mice

Whilst alterations in item binding in ASD have already been reported, the underlying mechanisms have not been described (Minor et al., 2023). Because sustained activity of dCA1 neurons during time intervals between the stimuli is necessary for long-term retention of their association(Al Abed et al., 2020b; Sellami et al., 2017), we examined dCA1 activity during acquisition in Cntnap2 KO mice. We recorded calcium transients during encoding and recall of trace fear conditioning, using fiber photometry in animals trained under a 20s or 40s-trace interval (Figure 1E). Both WT and Cntnap2 KO mice implanted with an optic fiber could acquire the tone-shock association, reaching high levels of freezing after three tone-shock pairings (Figure 1F, Figure S1E). In both T20 and T40 conditions, regardless of genotype, tone frequency significantly decreased across tone presentations, suggesting neural adaptation to successive tone exposure (Figure S1F, G). Cntnap2 KO mice trained under the 20s trace interval displayed no difference compared to WT in the frequency or amplitude of calcium transients during the tone and the trace interval (Figure S1F-H). In the 40s trace-conditioned group, while the frequency of the calcium transients did not differ between WT and Cntnap2 KO mice throughout conditioning (Figure S1G), their amplitude was significantly reduced both during the presentation of the second tone and consistently across the three trace intervals in Cntnap2 KO mice (Figure 1I). Overall, these results suggest a lack of synchronized neuronal activity during encoding, as previously reported (Simon and Wallace, 2016). During the tone test, as in Figure 1C, Cntnap2 KO mice displayed appropriate retention of the tone-shock association when conditioned under a 20s-trace interval (Figure S1H, I) while the ones trained under a 40s-trace interval failed to express a fear response to the tone. This indicates impaired retention of trace fear memory and temporal binding ability, relative to WT mice (Figure 2I). The findings also suggest that sustained activity of dCA1 neurons during encoding of the three trace intervals, which is higher in WT than in Cntnap2 KO mice, supports temporal binding and enables adequate retention of the association between the 40s-distant tone and shock despite their temporal separation, without affecting initial acquisition.

Cntnap2 KO mice display impaired relational/declarative memory.

A. Experimental design: Relational/Declarative memory task. ITI: Inter-trial interval; +: represent reward location within each pair of arms. B. Average correct answers for WT (black; n = 15) and Cntnap2 KO mice (purple; n = 14) for pairs A, B, and C during the 5 first days of acquisition (left panel; day (D); D4: Genotype: p=0.0149), and the last 5 days of acquisition (right panel; D-5: 5 days before they reach learning criterion; D-1: Genotype: p=0.0082). C. Comparison of the correct answers during the last day of training (D-1) averaged on pairs A and B, and during the flexibility test on the recombined pair AB (Genotype X Performance interaction: F1,27=12.639, p=0.0014; Test, Genotype: p=0.0239). Inset: D-1 vs. Test performance index on pairs AB (Genotype: p=0.0022). D. Comparison of the correct answers on pair C during the last day of training (D-1) and the flexibility test. Inset: D-1 vs. Test performance index for Pair C. E. Average performance on the new pair N during the flexibility test. F. Maximum lateralization index during stage 1 (absolute value of the average performance between pairs A and B (left baited) minus Pair C (right baited); Genotype: p=0.0153). G. Average decision latencies for correct (+) and incorrect (−) answers throughout training (Genotype X Response interaction: F1,31=7.960, p=0.0083). H. Average runtimes for correct (+) and incorrect (−) answers throughout training (Resp -, Genotype: p=0.0500). Dashed line illustrates chance level. Data shown at mean ± SEM. *: p<0.05; **: p<0.01; # shows genotype x performance interaction.

While the frequency of calcium transients remained unchanged in both T20 and T40 (Figure S1F, G), the amplitude of the peaks was higher during tone presentation in Cntnap2 KO mice conditioned under the 20s trace interval (Figure 2J). In contrast, transient amplitude was lower during both tone presentation and the subsequent trace period in Cntnap2 KO mice compared to WT mice (Figure 2K). These results further support the idea that the reduction in temporal binding capacity in Cntnap2 KO mice stems from a lack of sustained dCA1 activity during encoding and recall of the tone-shock association.

We previously demonstrated in WT mice that the activity of dCA1 principal neurons during the trace interval, but not during either the tone or the shock, is required for long-term retention, but not for encoding, of temporal associations (Sellami et al., 2017). To further confirm a causal link between sustained dCA1 activity during temporal gaps and 24h retention in Cntnap2 KO, we optogenetically activated principal neurons specifically during the 40s trace interval of the conditioning session, using ChannelRhodopsin (ChR2; Figure 1J), while leaving neuronal activity unaltered during the remainder of acquisition and the 24 h test. During the conditioning session, all groups displayed increased freezing across the three tone-shock pairings, indicative of efficient acquisition (Figure 1K, Figure S1J). 24h later, ChR2-expressing Cntnap2 KO mice demonstrated a specific response to the tone and trace compared to before and after the tone, unlike GFP-expressing Cntnap2 KO mice (Figure 1L, Figure S1K). These results demonstrate that dCA1 neuronal activity during trace intervals drives temporal binding in Cntnap2 KO mice.

Egocentric learning and inflexible long-term memory in Cntnap2 KO mice

Following the observation that Cntnap2 KO mice present reduced dCA1 activity associated with temporal binding capabilities, we examined potential declarative memory atypicalities in declarative memory. We employed a radial maze task that models the complexity of temporal binding-dependent relational/declarative memory (Sellami et al., 2018). This task assesses the characteristic flexibility of R/DM expression, which relies on the dCA1-dependent ability to relate temporally distant events (Sellami et al., 2017).

To do so, the R/DM task is divided into two stages: an acquisition phase and a flexibility test (Figure 2A). In the acquisition stage, mice are trained daily to learn the food-reward location within 3 separate pairs of adjacent arms. After reaching the learning criterion (>75% correct choices of the rewarded arm), the flexibility of the memory learnt during acquisition is probed by recombining two pairs, without changing the position of the reward (pairs A and B recombined into pair AB; Figure 2A). Cumulative evidence shows that flexibility relies on hippocampus-dependent relational representation of separate experiences formed for each pair during the acquisition phase (Al Abed et al., 2020b; Mingaud et al., 2007; Sellami et al., 2017). To impose a temporal binding demand on relational memory acquisition, the presentation of each pair is separated by a 20s inter-trial interval (Sellami et al., 2017).

Both wild-type (WT) and Cntnap2 KO mice were able to acquire the task (Figure 2B). Cntnap2 KO mice tended to require more days to reach criterion (Average number of days: WT: 8.06, KO: 10.86, ns, data in Figure S2) and reached higher performance levels at the end of training (Figure 2B, right panel).

In the flexibility test, their performance dropped significantly between the last training session and the flexibility test (stage 2 of the R/DM task), and Cntnap2 KO mice performed significantly worse than WT in the critical test of flexibility (Figure 2C - Recombined pair AB). In contrast, both groups performed similarly on the control pairs: the unchanged pair C, with high performance (Figure 2D), and the novel pair N, at chance levels (Figure 2E).

What could explain this lack of flexibility? We have previously shown that the flexibility component of R/DM depends on the learning strategy used during the acquisition stage: either allocentric spatial mapping (relational) or egocentric (procedural), the latter leading to lateralization. Indeed, when the use of spatial cues is impaired, as in cases of hippocampus damage or dysfunction, mice adopt a left-right egocentric strategy to memorize reward locations, leading to inflexible memory (Mingaud et al., 2007; Sellami et al., 2017). Therefore, the chance-level flexibility performance of Cntnap2 KO mice suggests a bias towards a procedural learning strategy, in line with the well-documented prominence of behavioral inflexibility (Uddin et al., 2015) and the use of egocentric learning strategies (Brunner et al., 2015; Ring et al., 2018) in ASD. This was confirmed by analyzing lateralization during acquisition by comparing performance on left-baited pairs (pairs A and B), vs. right-baited pair C (Methods, Figure 2F). Cntnap2 KO mice displayed significantly more lateralization than WT, meaning that they used an egocentric/procedural strategy to learn the position of the food reward rather than using spatial mapping, as observed in WT mice. We found no significant correlation between the use of egocentric strategy and the number of days to reach criterion (Data in Figure S2). Notably, performance during Day 4 of acquisition in WT, but not in Cntnap2 KO mice, was significantly negatively correlated with their use of lateralized strategy (r2 = −0.518, p=0.0471). Together, this further confirms that Cntnap2 KO mice employ a lateralized strategy to learn the task, which impacts both task performance (Figure 2B, day 4), and memory flexibility (Figure 2C).

Interestingly, Cntnap2 KO mice had shorter decision times than WT (Figure 2G), consistent with the well-demonstrated impulsivity observed in ASD (DSM-5 and American Psychiatric Association, 2013). In addition, WT mice demonstrated a longer time to enter a non-baited arm than a baited arm, suggesting that they evaluate reward location before entering the arm, likely relying on deliberation and precise spatial representation. In contrast, Cntnap2 KO mice showed similarly short latencies regardless of their choice, suggesting impulsive, automatic responses and impaired spatial representation of reward information. In addition, once a choice was made, their running time (from the central platform to the end of the arm) was longer than that of WT mice (Figure 2H), implying that the deliberative phase is shortened or absent in KO mice, while the slower run may reflect a compensatory adjustment following impulsive decisions.

Altogether, these findings show that Cntnap2 KO mice exhibit impaired flexible memory expression, modeling a relational/declarative memory deficit associated with preferential use of an egocentric/procedural learning strategy.

Potential shift towards striatum-centered activities in Cntnap2 KO mice

Each learning strategy is known to depend on partially distinct brain systems: hippocampus-dependent systems supporting allocentric spatial/relational learning, and striatum-based systems leading to egocentric/procedural learning (White and McDonald, 2002). To assess learning-induced activity in these different structures, we quantified the levels of the cellular activity marker cFos after day 3 of R/DM task acquisition in structures involved in memory (the dorsal hippocampus (dCA1, dCA2, dCA3, the dorsal dentate gyrus (dDG); the dorso-medial (DMS) and the dorso-lateral striatum (DLS); the prefrontal cortex (infralimbic (IL), prelimbic (PrL), and anterior cingular (ACC); Figure 3A).

Learning-induced activity and functional connectivity in Cntnap2 KO mice.

A. Experimental design: cFos levels were quantified 90min after the 3rd training session of the Relational/Declarative memory task. ITI: Inter-trial interval; + represents reward location within each pair of arms. B. number of cFos-positive neurons (cell density/mm2) in the dorsal dCA1 (WT naïve vs Trained: p=0.0491; Trained WT vs. KO: p=0.0129) with representative images of the cFos staining (Magenta) in dCA1, scale: 100µm, for naïve (light colors) and trained (dark colors) WT (grey/black) and Cntnap2 KO (purple) groups. C. number of cFos-positive neurons (cell density/mm2) in the Dorso-Medial Striatum (DMS; Naive WT vs. KO: p=0.0189; KO Naïve vs Trained: p=0.0085) with representative images of the cFos staining (Magenta) in DMS, scale: 200µm. D. Percentage of total activity: activity of each structure over to the total activity measured (percentage; Trained WT vs. KO: Hippocampus: p=0.00023; PFC: p=0.00139; Striatum: p=0.02895). E. Percentage of total activity: activity each subfield within the total network, in WT mice (naïve vs trained, left panel; dCA1: p=0.0382), in Cntnap2KO mice (Naïve vs trained, middle panel), and in trained mice (WT vs Cntnap2 KO mice, dCA1: p=0.0183; dCA2: p=0.0201; dCA3: p=0.0307; ACC: p=0.0378; DMS: p=0.0303; IL: p=0.0060; PrL: p=0.0476; right panel, same data as in left and middle panels). F. Pairwise between-structure correlation strength of cFos cell density. Data shown as mean ± SEM. *: p<0.05; **: p<0.01; # shows genotype x performance interaction.

As previously shown (Al Abed et al., 2020b; Sellami et al., 2017), we found a significant activation of dCA1 in trained WT mice compared to naïve WT mice, consistent with the use of an allocentric/spatial learning strategy (White and McDonald, 2002) (Figure 3B). In contrast, Cntnap2 KO mice showed no learning-induced dCA1 activation together with hypoactivation of dCA3 in both naïve and trained groups compared to WT mice (Figure S2A). This reduced hippocampal activation likely underlies the deficit in temporal binding and spatial learning capabilities in this task. Moreover, Cntnap2 KO mice displayed a significant training-induced activation of the DMS compared to their naïve controls, a pattern not observed in WT mice (Figure 3C). No other significant changes were detected in the DLS (Figure S2B) or in the prefrontal cortex regions (i.e.IL, PrL, ACC; Figure S2C). Together, this supports a shift toward DMS involvement in the egocentric learning strategy adopted by Cntnap2 KO mice.

To further dissect the balance between memory systems, we analyzed the relative contribution of each structure within the total activation of the system (Figure 3D-E). After training, we found a disproportionate representation of the PFC and striatum relative to the hippocampus in Cntnap2 KO mice compared to WT (Figure 3D).

Indeed, compared to WT naïve mice, WT trained mice exhibited an increase in the weight of the dCA1 in the network activity, likely supporting temporal binding and through ensuing relational organization of the 20s-distant pair presentations, the formation of flexible memory. In contrast, no such reorganization of network activity was observed in Cntnap2 KO mice with training (Figure 3E). Differences in the relative contribution of individual substructures between trained WT and Cntnap2 KO mice involved dCA1, dCA2, and dCA3, and favored the DMS, IL, and ACC. Overall, we observed compensatory activation of the frontal-striatal structures concomitant with hippocampal hypoactivation.

In addition to the learning-induced activity of individual structures, pairwise correlation of activity between structures enhances our understanding of network-wide synchronization. While naïve WT mice exhibited mainly positive correlations between the striatum, PFC, and hippocampus, with the exception of the dDG, which was negatively correlated with the PFC, naïve Cntnap2 KO mice displayed predominantly negative correlations (Figure 3F).

This indicates that, at baseline, interstructural relationships are dysregulated in Cntnap2 KO mice, potentially limiting the recruitment of a hippocampus-based learning strategy. Altogether, the analysis of learning-induced brain activity suggests a shift towards striatum-centered activity and connectivity in Cntnap2 KO mice, likely promoting an egocentric learning strategy and resulting in inflexible memory.

Discussion

This study establishes a novel mechanistic framework that builds on prior reports of hippocampal alterations in Cntnap2 KO mice (Paterno et al., 2021). It demonstrates that compromised CA1 hippocampal activity during encoding, disrupts temporal binding, which, by impairing the relational processing of temporally distant events, leads to deficits in retention of declarative memory. Moreover, hippocampal dysfunction in Cntnap2 KO mice appears to shift learning away from hippocampus-dependent strategies, and toward egocentric strategies.

Interplay between hippocampus and striatum in memory deficits in Cntnap2 KO mice

Atypicalities in declarative memory have been reported in ASD, with specific impairments in relating items (Minor et al., 2023). Here we demonstrate a reduction in dCA1-dependent temporal binding capacity in Cntnap2 KO mice, which can contribute to the declarative memory impairment by disrupting the relational organization of temporally distant events. Indeed, in the trace fear conditioning experiments, Cntnap2 KO mice showed efficient temporal binding with time intervals up to 20s, compared with 40s in WT mice, and this reduction was associated with a reduction of dCA1 activity during the encoding of the 40s-trace fear conditioning. Temporal integration of stimuli relies on hippocampal activity, in particular that of dCA1 (Buzsáki and Tingley, 2018; Chowdhury et al., 2005; Sellami et al., 2017; Tsien et al., 1996). Similarly, we found an impaired activation of the dCA1 in Cntnap2 KO mice during the acquisition of the R/DM task (Figure 3B).

The lack of training-induced dCA1 activation likely prevents fine-tuned temporal integration of distant events, leading to flexibility deficits (Chang and Gold, 2003; Etchamendy et al., 2003; Mingaud et al., 2007; Sellami et al., 2017). Moreover, analysis of broader memory systems’ activity during encoding revealed an aberrant recruitment of the DMS in the Cntnap2 KO mice, along with an altered pair-wise correlation between activation of related brain regions (Figure 3F). While WT mice exhibited predominantly positive correlations, Cntnap2 KO mice showed negative correlations among the DLS, prefrontal cortex and hippocampus. This pattern was also observed in naïve groups, suggesting that the bias towards lateralized learning strategy stems from baseline mistuning of functional connectivity between brain structures necessary for spatial learning in ASD.

This shift toward a striatum-centered system also implies a predominance of unconscious encoding, in which relations between items are acquired via stimulus–response associations, leading to inflexible, habit-like actions (Ferbinteanu, 2016). However, additional mechanisms may contribute to the phenotype. Cntnap2 KO mice exhibit other neural and behavioral alterations (Peñagarikano et al., 2011), including epileptic seizures and hyperlocomotion, suggesting that CA1 hypoactivity is likely only one of several factors influencing performance.

Decoding spatial and temporal maps in the Cntnap2 KO mouse model of autism

The CA1 region is recruited with other hippocampal areas to form and maintain spatial, temporal and social maps, which are mental relational representations of our environment (Barrientos and Tiznado, 2016). CA1 neurons contribute to encoding and recalling spatial information, enabling individuals to navigate and understand their surroundings. Specific “time cells” create a temporal framework that supports episodic memory, encoding the passage of time and the prediction of future events (Eichenbaum, 2014). Recent research also suggests that CA1 contributes to social cognition through so-called “social maps” which process and recall social information that is essential for effective social behavior and social dynamics (Zhang et al., 2022). Yet, little is known about how cognitive deficits evolve over time and their long-term effects in ASD. While the present study focused on dCA1 activation during acquisition to assess how dCA1 binds distinct stimuli into unified representations, future investigations directly examining the causal role of dCA1 activity in memory maintenance and retrieval (e.g. ArchT-mediated silencing during recall) will be essential for a comprehensive understanding of dCA1’s role in memory processes in ASD, as well as for exploring how specific maps are formed and maintained in this condition.

The role of dCA1 in spatial representations has been profusely documented. Consistently, we recently demonstrated the central role of dCA1 in the maladaptive contextual encoding in post-traumatic stress disorder (PTSD), i.e., the degradation of declarative memory surrounding a highly stressful event (Al Abed et al., 2020a; Brewin, 2011; DSM-5 and American Psychiatric Association, 2013). Interestingly, growing evidence reports the propensity to develop PTSD in ASD (Al Abed et al., 2024; Rumball et al., 2020), suggesting that reduced dCA1 activation may impact cognition beyond everyday learning. Together, these studies further underscore the hippocampus as a promising target to alleviate ASD pathophysiology.

To obtain a deeper understanding of the hippocampus computation underlying memory formation, we analyzed the correlation between neuronal dynamics and behavioral outcomes within the fear memory test. Fiber photometry recordings in the dCA1 showed that whilst WT mice displayed sustained hippocampal activity during trace interval (Figure 1I), hippocampus-dependent memory systems were not similarly activated in the Cntnap2 KO mice (Figure 3B, E). We found a marked decrease in the amplitude of calcium transients (Figure 1I). This profile was not observed when Cntnap2 KO mice were trained within their capacity (i.e. 20 s trace interval; Figure 1H).

Overall, these results suggest that the deficits in temporal binding and declarative memory originate from reduced hippocampal neuronal synchronization and dynamics. In the same line, a recent study reported during contextual fear conditioning, that neurons allocated to the memory trace displayed freezing-specific activity compared to non-allocated neurons (Mocle et al., 2024). While our study design does not provide these details, the lack of active cells in the dCA1 during acquisition of the relational/declarative memory task could imply memory trace allocation to the striatum rather than the hippocampus (Rashid et al., 2016). Further experiments are needed to determine whether atypical engram allocation is linked to learning strategies in ASD.

Aged-like performance in Cntnap2 KO mice suggests early cognitive decline

In this study, we demonstrate the disruption in hippocampal and CA1 functioning in the Cntnap2 KO mice, a hallmark of memory defects observed during aging (Etchamendy et al., 2012; Leal and Yassa, 2015). Consistent with reports suggesting that people with autism are at risk of premature cognitive aging (Wang et al., 2024), we have found that young adult Cntnap2 KO mice exhibit a memory profile similar to those of aged WT mice (Al Abed et al., 2020b; Marighetto et al., 2011; Sellami et al., 2017). Compared to young WT mice, both groups showed flexibility impairments in the relational/declarative radial maze task, and both exhibited a striatum-biased learning strategy. Yet, the underlying mechanisms driving these deficits remain unknown.

We previously demonstrated that dCA1 activity during the temporal gaps is necessary to link separate pairs into a relational/flexible memory (Al Abed et al., 2020b; Sellami et al., 2017) and that targeted dCA1 optogenetic activation during these intervals increases temporal binding capacity, rescuing relational/declarative memory formation. Given that CA1 neurons are less synchronized in ASD, targeted dCA1 activation could alleviate memory deficits. We propose that fine-tuning dCA1 activity may normalize memory system balance, offering a novel therapeutic approach for cognitive impairments in ASD. Further delineation of associated connectomes (Hollunder et al., 2024) could provide insights into targeted brain circuit therapies.

Autism is a highly heterogeneous condition, with individuals presenting a wide range of atypical cognitive abilities. Consequently, the extent and nature of memory deficits vary considerably across the spectrum. Research has yet to fully elucidate how these deficits manifest, particularly in relation to other cognitive domains such as attention, executive functioning, and language. Sex differences are another factor to consider (Van Wijngaarden-Cremers et al., 2014). While this study included both male and female mice, we did not observe sex-specific effects, suggesting that the behavioral and cognitive phenotypes examined in the Cntnap2 KO mouse model are largely independent of sex. This is consistent with previous studies reporting no robust sex differences (Al Abed et al., 2024).

A better understanding of these individual differences is necessary to create more personalized and effective interventions. As such, further studies will have to assess these aspects in different models to fully comprehend the common cognitive mechanisms at play in autism.

Methods and material

Mice

3 to 5-month-old naive mice were collectively housed in standard Makrolon cages, in a temperature- and humidity-controlled room under a 12h light/dark cycle (lights on at 07:00), with ad libitum access to food and water. We used both males and females to minimize the number of mice produced in this study. Cntnap2+/+ (WT) and Cntnap2−/− (KO) mice (Jackson ID: #017482 (Peñagarikano et al., 2011)) were non-littermate, though both lines were originally derived from breeders of the same colony and were used for behavior and histological analyses. To minimize potential differences between strains, we regularly inter-crossed the wildtype and Cntnap2 KO colonies (i.e., every year) to generate heterozygous mice. These heterozygotes were then bred to produce either WT or KO mice for experiments. Throughout the study, we used non-littermate mice originating from the same breeders. To ensure that observed differences between WT and Cntnap2 KO were not influenced by this breeding strategy, we also assessed trace fear conditioning capacities in littermate mice and confirmed our findings using Het × Het breeding experiments (Figure S1).

All mice were bred in the same room, and maintained under identical husbandry and rearing conditions, including cage content, bedding, diet, and light/dark cycle. All experiments took place during the light phase. We replicated the behavioral experiments in two to four different batches. Every effort was made to minimize the number of animals used and their suffering. All procedures were conducted in accordance with the European Directive for the care and use of laboratory animals (2010-63-EU) and the animals care guidelines issued by the animal experimental committee of Bordeaux University (CCEA50, agreement number A33-063-099; authorization N°21248), and from the Australian National University Animal Experimentation Ethics Committee (protocol numbers A2018/66, A2020/26, A2021/43, and A2024/362).

Trace fear conditioning task (TFC)

Apparatus

Fear conditioning behavior was performed in a Plexiglas conditioning chamber (20 x 20 x 30 cm), in a brightness of 100 lux, given access to the different visual-spatial cues in the experimental room. The floor of the chamber consisted of stainless-steel rods connected to a shock generator (Imetronic, Pessac, France).

Trace fear conditioning procedure

The box was cleaned with 70% ethanol before each trial. During conditioning, each animal received 3 pairings of a tone (70 dB, 1 kHz, 30 s) and a footshock (0.3 mA, 50 Hz, 1 s). The two stimuli were separated by a trace interval of 5, 20, or 40 s, depending on the group. Each tone-shock pairing was separated by a 1min-interval.

All mice were submitted 24 hours later to the Tone retention test during which they were re-exposed to the tone alone in a dark and modified chamber [circular opaque white chamber cleaned with a 1% acetic acid solution; 2min pre-tone, 2min tone and 2min post-tone]. A context test was also performed 2 hours later, during which mice were re-exposed to the conditioning chamber alone for 6 min.

Animals were continuously recorded for off-line scoring of freezing, by an experimenter ignorant of the experimental condition/group. Tone test was scored by 2 independent scorers. Freezing is defined as a lack of all movement except for respiratory-related movements. Acquisition of the TFC was measured by the progression of freezing during the tone across the 3 tone deliveries. 24h-retention of the trace association was measured during the tone test by the evolution of percentage of freezing time. Conditioned fear response to the tone was evaluated by calculating a normalized ratio between the tone and trace periods, versus outside these periods (i.e., Pre-Tone and Post-Tone): [(Tone+Trace)/2 - (pre-Tone+post-Tone)/2] / [(Tone+Trace)/2 + (pre-Tone+post-Tone)/2].

Radial-maze task of Relational/Declarative memory (R/DM)

Apparatus

We used an open 8-arm radial maze made of grey Plexiglas, automatized by videotracking (IMETRONIC – Pessac, France). The diameter of the central platform is 30 cm, and the arms are 55 cm long by 10 cm large. Each arm is equipped with a door at its entrance and a food-pellet delivery system at its end. The doors are individually controlled (raised up or dropped down) by the computerized system which also controls pellets availability in the food tray at the end of each arm individually, according to the task. The maze is placed in an empty room containing visual cues to enable spatial discrimination.

Behavioral Procedure

During the entire procedure, animals were submitted to one daily session. Prior to memory testing, animals were habituated to the apparatus over a period of two days during which animals were allowed to move freely in the radial maze. To complete the session, mice had to visit each arm until its end. Choice accuracy was measured as the percentage of correct responses for each pair.

Stage 1: Acquisition of 3-pair discriminations

The acquisition task consisted in learning the position of the food within the maze. Indeed, each animal was assigned three adjacent pairs of arms (pairs A, B and C). In each pair, only one arm was baited as a food reward. The experiment is designed in a way that the left arms of pairs A and B are baited while the right arm of pair C is. In each trial, mice were given access to a pair (either pairs A, B, or C). A choice was made when the subject reached the food well of an arm; this also closed the door of the non-chosen arm. The trial was finished as soon as the animal returned to the central platform. The subject was then confined to the central platform for 20s before the next discrimination trial began: this constituted the inter-pair interval (ITI). Each daily session consisted of 20 consecutive trials comprising alternate presentations of pairs A, B, and C according to a pseudo-random sequence. A mouse was considered to reach criterion when its overall choice accuracy was at or above 75% over two consecutive sessions given that performance in each of the three discrimination choices were at least 62% correct. The flexibility test task began the following day when criterion performance was reached. Lateralization index was quantified by using the absolute value of the subtraction of the performance on the left-baited pairs vs. the right-baited pairs. Lat = |((Pair A + Pair B) /2) - Pair C|. We then used the maximum lateralization over 3 consecutive days.

Stage 2: Flexibility probe of DM

In the test task, the position of the food in the maze did not change, but characteristic flexibility of DM expression was assessed by changing the way of presenting the arms. Indeed, in place of the pairs A and B, a pair AB was submitted to the mouse. The AB pair consisted in the combination of the two adjacent arms of the pairs A and B. It was the critical test of flexibility. Two other pairs were used as control: the pair C that remained unchanged (“unchanged learnt control”, and pair N (=new) made of the two arms non-used in acquisition (“unlearnt control”).

Fiber photometry

Surgery

Mice were injected unilaterally 4 weeks before behavioral experiments with an Adeno-Associated Virus (AAV) GCaMP6f (AAV1-hsyn-jGCaMP6f-WPRE, Addgene). We used glass pipettes (tip diameter 25-35 µm) connected to a picospritzer (Parker Hannifin Corporation) into the right side of the dorsal CA1 of the hippocampus (0.1 µl/site; AP −1.8 mm; L +1.35 mm; DV −1.4 mm). Mice were then implanted with unilateral optic fiber implant (diameter: 200µm; numerical aperture: 0.39; flat tip; Neurophotometrics). Implants were fixed to the skull with Super-Bond dental cement (Sun Medical, Shiga, Japan). Mice were perfused after experiments to confirm correct placements of fibers. Viral injections targeted the pyramidal cell layer of the dCA1.

Recordings

Data were recorded throughout the entire fear conditioning paradigmusing a Neurophotometrics FP3002 system (Neurophotometrics, San Diego, CA). Briefly, 470 nm and 415 nm LED light was bandpass filtered, reflected by a dichroic mirror, and focused onto a multi-branch patch cord (Neurophotometrics, San Diego, CA) by a x20 objective lens. Alternating pulses of 470 and 415 nm light (∼50 µW) were delivered at 60 Hz, and photometry signals were time-locked to behavior using a Bonsai workflow. Signals were analyzed using the Synaptech suite cloud service (MetaCell) and custom-made Python scripts (Code available at https://github.com/elisecr/fiberphotometry_2024). Through the Synaptech platform, 470 and 415 signals were deinterleaved. For all experiments, the exponential delay period (the first 10 seconds(Hon et al., 2025)) of the recording was removed. To correct for photobleaching, isosbestic (415nm) signals were fit to a biexponential curve to model the baseline signal with photobleaching. The fitted biexponential curve was then scaled to match the 470nm signal using non-negative robust linear regress and subtracted from the 470 trace to correct for baseline drift. Traces were then Z-scored.

Experimental traces

Z scored data were analyzed by custom python scripts to extract calcium fluorescence traces for each experiment. For the trace fear conditioning experiments, these data were aligned to files generated by the poly fear software (Imetronic, France) “.dat” files to align the beginning of the tone and shock events to the closest millisecond in the Z scored data. The timing of the tones and trace periods were extracted as described above to accurately align the data.

Peak finding

The threshold for peak analysis definition was set at 1.0. Peak analysis was conducted using a custom-built Python script to calculate peak frequency and amplitude. Code available at https://github.com/elisecr/fiberphotometry-analysis

Optogenetic Manipulation of dCA1

Surgery

Mice were injected bilaterally 5 weeks prior to behavioral experiments with an Adeno-Associated Virus (AAV) containing the excitatory opsin channelrhodopsin targeting the pyramidal cells of dCA1 (pAAV-CaMKIIa-hChR2(H134R)-mCherry (#26975), Addgene). We used glass pipettes (tip diameter 25-35 µm) connected to a picospritzer (Parker Hannifin Corporation) into the dorsal CA1 of the hippocampus (0.2 µl/site; AP −1.8 mm; L +1.3 mm; DV −1.4 mm and AP −2.5 mm; L +2 mm; DV −1.4 mm). Mice were then implanted with bilateral optic fiber implant (AP −1.8 mm; L +1.3 mm; DV −1.3 mm; diameter: 200µm; numerical aperture: 0.39; flat tip; Neurophotometrics). Implants were fixed to the skull with Super-Bond dental cement (Sun Medical, Shiga, Japan). Mice were perfused after experiments to confirm correct placements of fibers. Viral injections targeted the pyramidal cell layer of the dCA1. Optogenetic manipulation during trace fear conditioning: Mice were conditioned under a 40s trace interval condition, while dCA1 pyramidal neurons were activated during the 40 seconds of trace interval only. The light (≈3 mW per implanted fiber) was bilaterally conducted from the laser (Neurophotometrics, 450nm; 5 Hz; 5 ms laser on, 195 ms laser off) to the mice via two fiber-optic patch cords (diameter, 200 μm; Doric Lenses). The tone test was performed 24h later, mice were tethered to the patch chord, without optogenetic manipulation.

Open field

Mice were placed in a 38.5cm diameter circular arena and allowed free exploration for 5 minutes. These movements were recorded, and the mean velocity, distance travelled, and resting time were tracked and quantified using a custom MATLAB (Mathworks) code as in (Parkinson et al., 2024).

Immunohistochemistry

Naïve (homecage) and trained (90 min after the flexibility test on Day 3 for c-Fos analysis) animals were perfused transcardially with 0.01M phosphate buffered saline (PBS) to eliminate blood and extraneous material, followed by 4% paraformaldehyde (PFA). Brains were postfixed for 36h in PFA. Tissues were sectioned at 40 µm using a Leica 1000S vibratome and kept in a cryoprotective ethylene glycol solution at −20°C until processed for immunofluorescence. Sections were first washed and permeabilized in PBS-Triton 0.25%, then non-specific binding sites were blocked by immersing the tissue in 10% normal donkey serum, 2% bovine serum albumin in PBS-Triton 0.25% during 2h. Tissues were then stained using the primary antibodies overnight: mouse anti-c-Fos (1:1000; Santa Cruz). After 3x 15 min washes, we added anti-rabbit Alexa 488 (1:200; Life Technologies) secondary antibodies for 2h. After 3x 15 min washes slices were stained during 5 min with DAPI (5µM; Sigma), mounted on Livingstone slides then covered with Mowiol (Sigma) and coverslip (Thermofisher). c-Fos, staining from 2-3 slices par animals were imaged using a Nikon Axioscan confocal fluorescent microscope (20x objective). Stained sections of WT and mutant mice were imaged during the same imaging session. Immunofluorescence signals were quantified using the quPath software with routine particle analysis procedures (size= 30-500; circularity=0.30-1.00), to obtain cellular masks, divided by the area to obtain cell density per mm2.

Statistics

Data are presented as mean ± SEM. Statistical analyses were performed using GraphPad prism or StatView software for 2-way ANOVA, followed by Bonferroni post-hoc test or Student’s t test, when appropriate. Normality of the data was confirmed using the Kolmogorov–Smirnov test. Statistical significance was considered at p<0.05.

eLife Assessment

This valuable study uses the Cntnap2 mouse model of autism to investigate how reduced activity in the dorsal CA1 region of the hippocampus contributes to impaired temporal binding and memory flexibility. The authors combined trace conditioning and radial maze assays with fibre photometry, optogenetic rescue, and brain-wide c-Fos mapping to provide solid evidence that dorsal CA1 hypoactivity limits the retention of temporally separated associations and biases learning towards less flexible strategies. The work advances understanding of hippocampal contributions to cognitive alterations associated with autism and will be of broad interest to researchers studying memory, hippocampal function, and neurodevelopmental disorders.

Reviewer #1 (Public review):

Summary:

The uniqueness of this paper is the study of the formation of temporal binding-dependent memories in the cntnap2 mouse, a long-standing mouse model of autism that has been used to test therapeutic modalities.

Strengths:

I liked the combination of optical recordings and interventions and the backup of primary observations with control experiments.

Weaknesses:

(1) Fiber photometry recordings are too coarse to give salient clues to the underlying mechanism.

(2) Are perturbed pyramidal cells causally responsible for the altered trace? What can be concluded about the possible role of inhibitory interneurons as potential drivers? The observations focus on abnormal regional activity as observed with fiber photometry and manipulated by optogenetics. The authors should state clearly the limits of their conclusions.

(3) I found the "trace" nomenclature confusing. "....in which mice are required to memorize the association between a tone (Conditioned Stimulus) and a mild electric foot-shock (Unconditioned Stimulus), separated by a time interval called Trace (Sellami et al., 2017)." It seems that the conceptual model invokes the creation of an [eligibility] trace, characterized by its progressive disappearance over time. It may be a convention in the field or a matter of language, but it seems perverse to use "trace" to label the time interval rather than the entity that is decaying. If this is an accepted convention going back to Howard Eichenbaum, the authors should cite the paper that first introduced the convention.

(4) I would advocate for the addition of some discussion points for the authors to consider.

a) Is the retention of activity in CA1 related to phenomena at the cellular or subcellular level in CA1 pyramidal cells? I'm thinking of dendritic, delayed, and stochastic CaMKII activation (DDSC) as defined by Yasuda's group or short-term and associative plasticity of calcium dynamics (STAPCD) as delineated by Caya-Bissonette and Beique.

b) Was the optogenetic intervention ever administered in a delayed fashion, capitalizing on the temporal advantages of optogenetics to probe dynamics?

c) Is the newfound reliance on corticostriatal pathways something more than compensation at the behavioral level? Could it be driven in part by the ASD-related genetic changes?

Reviewer #2 (Public review):

The authors investigate the contribution of dorsal CA1 hippocampal dysfunction to cognitive impairments in the Cntnap2 knockout mouse model of autism spectrum disorder. Building on previous evidence implicating the hippocampus in episodic and relational memory processes, they combine trace fear conditioning, fiber photometry, optogenetic manipulation, a relational/declarative memory radial maze task, and cFos mapping to test whether altered CA1 function contributes to deficits in temporal binding and memory flexibility.

The study has several important strengths. First, the work addresses a relatively understudied aspect of autism-related cognition, namely hippocampal-dependent memory processes, whereas much of the literature has focused on social behavior, cortical circuits, or striatal dysfunction. Second, the authors employ multiple complementary approaches that converge on a coherent mechanistic hypothesis. The behavioral data demonstrate a reduced ability of Cntnap2 knockout mice to retain associations across long temporal gaps. Fiber photometry recordings reveal reduced dorsal CA1 activity during conditions that challenge temporal binding, and optogenetic activation of dorsal CA1 neurons during the trace interval is sufficient to rescue memory performance. Together, these findings provide strong support for a causal contribution of dorsal CA1 activity to temporal binding deficits in this model.

The second major strength of the manuscript is the extension of these findings to a more complex hippocampus-dependent memory task. The radial maze experiments indicate that Cntnap2 knockout mice show impaired memory flexibility and a greater reliance on egocentric learning strategies. The accompanying cFos analyses suggest altered recruitment of hippocampal and striatal networks during learning, providing a systems-level framework that may explain the observed behavioral phenotype.

Overall, the main conclusions regarding impaired temporal binding and reduced dorsal CA1 engagement are well supported by the data. The optogenetic rescue experiments are particularly compelling because they move beyond correlation and directly test causality. The manuscript therefore makes a meaningful contribution to our understanding of how hippocampal dysfunction may contribute to cognitive abnormalities associated with autism.

Weaknesses:

Some conclusions are necessarily more inferential than others. In particular, the interpretation that the observed behavioral phenotype reflects a broader shift from hippocampal-dependent declarative memory toward striatum-dependent procedural learning is supported primarily by cFos activity patterns and behavioral strategy measures. While the data are consistent with this interpretation, they do not directly demonstrate a causal reorganization of memory systems. Similarly, although the findings identify a mechanism in the Cntnap2 model, caution is warranted when extrapolating these conclusions to autism spectrum disorder more broadly; but I believe this caution is addressed in the discussion.

Despite these limitations, the study presents a coherent and well-executed body of work that provides novel mechanistic insight into hippocampal contributions to cognitive dysfunction in a widely used autism model. The findings should be of considerable interest to researchers studying hippocampal function, memory systems, and neurodevelopmental disorders.

Reviewer #3 (Public review):

Summary:

The manuscript evaluated behavioral phenotypes in the Cntnap2 knockout mouse using two behavioral paradigms: trace fear conditioning and a radial maze task. The trace fear conditioning training is normal, but memory generalization is impaired. The inflexibility is suggested to be related to low activity in dCA1 neurons, which can be rescued by ChR2. The radial maze task data suggested a similar conclusion. Brain-wide cFos mapping indicated impairments in the Cntnap2 knockout mouse. The brain-wide cFos mapping does not show direct correlations with Cntnap2, limiting the interpretation of these data in the context of this paper.

Strengths:

The behavior data are solid.

Weaknesses:

The underlying mechanism is not fully investigated.

Major points：

(1) The authors should thoroughly check their manuscript as there are many typos in the current version that affect the readability.

(2) In trace fear conditioning, the tone test impairment can be rescued by ChR2. Have the authors tried rescue experiments with Cntnap2? Rescue experiments in the radial maze task are also essential, either with ChR2 or Cntnap2.

(3) The quality of the cFos example image in Figure 3 is too low. The authors should also provide example images for the other brain regions in the supplementary data, if possible.

(4) The causal link between the brain-wide cFos mapping and the Cntnap2 knockout is weak. How to explain the increase of cFos cell densities in some brain regions, but the decrease in others?

Author response:

The following is the authors’ response to the original reviews.

Public Reviews:

Reviewer #1 (Public review):

Summary:

The uniqueness of this paper is the study of the formation of temporal binding-dependent memories in the cntnap2 mouse, a long-standing mouse model of autism that has been used to test therapeutic modalities.

Strengths:

I liked the combination of optical recordings and interventions and the backup of primary observations with control experiments.

Weaknesses:

(1) Fiber photometry recordings are too coarse to give salient clues to the underlying mechanism.

We acknowledge that fiber photometry provides population-level measurements and does not resolve the activity of individual neurons or synaptic mechanisms. Our aim was to identify alterations in the activity of defined neuronal populations during behaviour rather than to delineate the underlying cellular mechanisms. We agree that future studies employing higher-resolution approaches, such as two-photon calcium imaging, in vivo electrophysiology, or single-cell recordings, would provide important mechanistic insights into the circuit changes underlying the observed activity patterns.

(2) Are perturbed pyramidal cells causally responsible for the altered trace? What can be concluded about the possible role of inhibitory interneurons as potential drivers? The observations focus on abnormal regional activity as observed with fiber photometry and manipulated by optogenetics. The authors should state clearly the limits of their conclusions.

Our optogenetic experiments demonstrate a causal role for the targeted neuronal population in modulating the observed activity and behavioural phenotype. However, we do not conclude that this population is solely responsible for generating the altered fiber photometry signal, nor do we infer that it represents the exclusive driver of the underlying circuit dysfunction. Rather, our findings demonstrate that manipulating this population is sufficient to alter circuit activity, while acknowledging that the recorded signals likely reflect interactions between multiple neuronal populations. We have revised the manuscript to make these distinctions clearer.

(3) I found the "trace" nomenclature confusing. "....in which mice are required to memorize the association between a tone (Conditioned Stimulus) and a mild electric foot-shock (Unconditioned Stimulus), separated by a time interval called Trace (Sellami et al., 2017)." It seems that the conceptual model invokes the creation of an [eligibility] trace, characterized by its progressive disappearance over time. It may be a convention in the field or a matter of language, but it seems perverse to use "trace" to label the time interval rather than the entity that is decaying. If this is an accepted convention going back to Howard Eichenbaum, the authors should cite the paper that first introduced the convention.

We thank the reviewer for this comment. This is indeed an established convention in the classical/Pavlovian conditioning literature rather than terminology specific to our study or to Sellami et al. (2017). The term "trace conditioning" was coined by Pavlov (1927), who used "trace interval" to designate the empty period separating CS offset from US onset, precisely because — as the reviewer intuits — successful conditioning across this gap requires the organism to maintain a memory trace of the CS. In other words, the interval is named for the cognitive/neural entity (the decaying CS trace) that must be sustained across it, not because the interval itself is thought to be a physical or decaying object. So "trace interval" is shorthand for "the interval across which a trace must be maintained," analogous to how "delay conditioning" refers to a paradigm named for a temporal property of the procedure rather than the mechanism per se. We have added a citation to Pavlov (1927) at first use of the term, and clarified the phrasing to make the etymology explicit, as suggested.

(4) I would advocate for the addition of some discussion points for the authors to consider.

(a) Is the retention of activity in CA1 related to phenomena at the cellular or subcellular level in CA1 pyramidal cells? I'm thinking of dendritic, delayed, and stochastic CaMKII activation (DDSC) as defined by Yasuda's group or short-term and associative plasticity of calcium dynamics (STAPCD) as delineated by Caya-Bissonette and Beique.

We thank the reviewer for this insightful suggestion. Our study was designed to investigate network-level dynamics, and the approaches used do not allow us to determine whether the retained activity in CA1 arises from intracellular mechanisms, such as dendritic calcium dynamics or CaMKII-dependent signalling (including mechanisms such as DDSC or STAPCD), recurrent circuit interactions, or a combination of both. We therefore cannot directly assess the contribution of these cellular and subcellular processes. We have now added a paragraph to the Discussion acknowledging that persistent dendritic calcium signalling and CaMKII-dependent plasticity are plausible contributors to sustained CA1 activity and represent an important avenue for future investigation.

(b) Was the optogenetic intervention ever administered in a delayed fashion, capitalizing on the temporal advantages of optogenetics to probe dynamics?

This is indeed an important control. While we did not perform this intervention in the current study, this control was part of our seminal study demonstrating the causal role of the dCA1 in temporal binding (Sellami et al., PNAS, 2017, Fig. 1E doi: https://doi.org/10.1073/pnas.161965711). We showed that aged mice had reduced temporal binding capacity, associated with decreased dCA1 activity. We successfully rescued temporal binding capacity in aged mice through ChR2-induced activation of dCA1 pyramidal neurons during the trace interval, but not outside of the trace interval. Given the striking similarity between the temporal binding deficits observed in aged mice and those reported here in Cntnap2 KO mice, we did not repeat this previously established control in the present study. To acknowledge this point, we have also added a statement to the Discussion noting that confirming the temporal specificity of dCA1 optogenetic manipulation in the Cntnap2 KO model will be an important direction for future studies.

(c) Is the newfound reliance on corticostriatal pathways something more than compensation at the behavioral level? Could it be driven in part by the ASD-related genetic changes?

We agree that the increased reliance on corticostriatal pathways could reflect both compensatory recruitment at the behavioral level and a direct consequence of ASD-related genetic alterations affecting circuit development and function. Our current data demonstrate a shift in circuit engagement but do not allow us to distinguish whether this represents an adaptive compensation or a primary consequence of Cntnap2 deficiency. However, given that ASD-associated mutations can alter the development, connectivity, and plasticity of corticostriatal circuits, it is possible that the observed changes reflect intrinsic circuit reorganization rather than solely a compensatory behavioral strategy. We have revised the Discussion to acknowledge this possibility and to clarify that altered corticostriatal recruitment may represent a direct consequence of the genetic disruption that subsequently shapes behavioral strategies.

Reviewer #2 (Public review):

The authors investigate the contribution of dorsal CA1 hippocampal dysfunction to cognitive impairments in the Cntnap2 knockout mouse model of autism spectrum disorder. Building on previous evidence implicating the hippocampus in episodic and relational memory processes, they combine trace fear conditioning, fiber photometry, optogenetic manipulation, a relational/declarative memory radial maze task, and cFos mapping to test whether altered CA1 function contributes to deficits in temporal binding and memory flexibility.

The study has several important strengths. First, the work addresses a relatively understudied aspect of autism-related cognition, namely hippocampal-dependent memory processes, whereas much of the literature has focused on social behavior, cortical circuits, or striatal dysfunction. Second, the authors employ multiple complementary approaches that converge on a coherent mechanistic hypothesis. The behavioral data demonstrate a reduced ability of Cntnap2 knockout mice to retain associations across long temporal gaps. Fiber photometry recordings reveal reduced dorsal CA1 activity during conditions that challenge temporal binding, and optogenetic activation of dorsal CA1 neurons during the trace interval is sufficient to rescue memory performance. Together, these findings provide strong support for a causal contribution of dorsal CA1 activity to temporal binding deficits in this model.

The second major strength of the manuscript is the extension of these findings to a more complex hippocampus-dependent memory task. The radial maze experiments indicate that Cntnap2 knockout mice show impaired memory flexibility and a greater reliance on egocentric learning strategies. The accompanying cFos analyses suggest altered recruitment of hippocampal and striatal networks during learning, providing a systems-level framework that may explain the observed behavioral phenotype.

Overall, the main conclusions regarding impaired temporal binding and reduced dorsal CA1 engagement are well supported by the data. The optogenetic rescue experiments are particularly compelling because they move beyond correlation and directly test causality. The manuscript therefore makes a meaningful contribution to our understanding of how hippocampal dysfunction may contribute to cognitive abnormalities associated with autism.

Weaknesses:

Some conclusions are necessarily more inferential than others. In particular, the interpretation that the observed behavioral phenotype reflects a broader shift from hippocampal-dependent declarative memory toward striatum-dependent procedural learning is supported primarily by cFos activity patterns and behavioral strategy measures. While the data are consistent with this interpretation, they do not directly demonstrate a causal reorganization of memory systems. Similarly, although the findings identify a mechanism in the Cntnap2 model, caution is warranted when extrapolating these conclusions to autism spectrum disorder more broadly; but I believe this caution is addressed in the discussion.

We agree that our data do not directly demonstrate a causal reorganization of memory systems with cFOS activity patterns. Our intention was to propose that the behavioural strategy together with the brain-wide cFos activation patterns are consistent with a shift in the relative engagement of hippocampal- and striatal-dependent networks. We have revised the manuscript to moderate our interpretation throughout, replacing causal language with wording that reflects an association between the observed behavioural changes and altered recruitment of these memory-related circuits.

Despite these limitations, the study presents a coherent and well-executed body of work that provides novel mechanistic insight into hippocampal contributions to cognitive dysfunction in a widely used autism model. The findings should be of considerable interest to researchers studying hippocampal function, memory systems, and neurodevelopmental disorders.

Reviewer #3 (Public review):

Summary:

The manuscript evaluated behavioral phenotypes in the Cntnap2 knockout mouse using two behavioral paradigms: trace fear conditioning and a radial maze task. The trace fear conditioning training is normal, but memory generalization is impaired. The inflexibility is suggested to be related to low activity in dCA1 neurons, which can be rescued by ChR2. The radial maze task data suggested a similar conclusion. Brain-wide cFos mapping indicated impairments in the Cntnap2 knockout mouse. The brain-wide cFos mapping does not show direct correlations with Cntnap2, limiting the interpretation of these data in the context of this paper.

We agree that brain-wide cFos mapping does not directly identify the molecular or cellular mechanisms by which Cntnap2 deficiency alters circuit function. Rather, we use cFos as a functional readout of network recruitment during behaviour. Our interpretation is therefore limited to identifying differences in activity patterns associated with the behavioural phenotype, rather than establishing direct mechanistic links to Cntnap2 function. We have clarified this point in the revised manuscript and moderated the text accordingly.

Strengths:

The behavior data are solid.

Weaknesses:

The underlying mechanism is not fully investigated.

Major points：

(1) The authors should thoroughly check their manuscript as there are many typos in the current version that affect the readability.

(2) In trace fear conditioning, the tone test impairment can be rescued by ChR2. Have the authors tried rescue experiments with Cntnap2? Rescue experiments in the radial maze task are also essential, either with ChR2 or Cntnap2.

We agree that rescue experiments in the radial maze would provide additional mechanistic insight by testing whether restoring dCA1 activity optogenetically in Cntnap2 KO mice is sufficient to rescue declarative memory flexibility. However, we have already established the causal role of dCA1 in temporal binding and relational/declarative memory in the radial maze task in our earlier work (Sellami et al., PNAS, 2017, Figure 2E). Indeed, optogenetic inhibition of dCA1 in the radial maze task, specifically during the inter-trial interval, prevented flexible relational/declarative memory formation. In the present study, we confirmed the causal contribution of dCA1 activity to temporal binding in Cntnap2 KO mice in the trace fear conditioning paradigm (Figure 1J-L). Therefore, repeating the optogenetic experiment in the radial maze would provide only limited additional information, while requiring dedicated experimental cohorts and additional validation. Moreover, as the team is based in 2 different institutions and countries (France and Australia), we unfortunately do not have the capacity to perform these experiments.

Regarding a rescue with the Cntnap2 protein, here, we used the Cntnap2 KO as a well-accepted model of autism spectrum disorder (ASD), rather than understanding the role of this protein in temporal binding capacity and memory formation. The present study was hence designed to determine how dCA1 activity relates to temporal binding and memory performance in ASD, building on the causal evidence established in our previous work. We therefore consider Cntnap2-specific rescue experiments an important follow-up to further test the sufficiency of dCA1 activation, rather than an essential experiment for establishing the conclusions of the present study.

We have highlighted these points as an important direction for future investigations and revised the manuscript to better distinguish our findings from this additional mechanistic question.

(3) The quality of the cFos example image in Figure 3 is too low. The authors should also provide example images for the other brain regions in the supplementary data, if possible.

The low quality of the cFos image in Figure 3 resulted from a formatting issue during figure uploading. This has now been corrected, and a higher-resolution image has been included in the revised manuscript. We have also added representative images for the other brain regions analyzed to the Supplementary Information, as requested.

(4) The causal link between the brain-wide cFos mapping and the Cntnap2 knockout is weak. How to explain the increase of cFos cell densities in some brain regions, but the decrease in others?

We agree that cFos mapping cannot explain the mechanisms underlying the regional increases and decreases in activity. We interpret these bidirectional changes as reflecting differential recruitment of distributed brain networks rather than direct effects of Cntnap2 deficiency on individual brain regions, and have clarified this in the revised manuscript.

Recommendations for the authors:

Reviewer #1 (Recommendations for the authors):

(1) Improve clarity of the Figure layout: Figure 1, panel I legend. I think the identification of experimental groups is in the wrong place. It really applies to panels K, N, and L, not to panel I. This is important because the genetic stimulation in panels K, N, and L is the key experiment in this whole figure.

We agree that the placement of the experimental group identification in the original figure legend could be confusing. We have revised the figure layout and legend so that the description of the experimental groups is now associated with panels K and L, where the optogenetic stimulation experiments are presented, thereby improving the clarity of the figure.

(2) Fix Figure references. "This indicates impaired retention of trace fear memory and temporal binding ability, relative to WT mice (Figure 2I)....: [there is no Fig. 2I, so I'm guessing you mean Fig. 1I; subsequent Figure references also seem to confuse Fig. 1 and 2].

We thank the reviewer for identifying these errors. We have carefully checked and corrected all figure citations throughout the manuscript, including the reference to Figure 1I-L in this section, to ensure that each citation now refers to the appropriate panel.

(3) Missing words? In both T20 and T40 conditions, regardless of genotype, [the response to] tone frequency significantly decreased.

We thank the reviewer for spotting this omission. The sentence has been corrected to read: "In both T20 and T40 conditions, regardless of genotype, the response to tone frequency significantly decreased."

(4) Jargon that should be declared at an appropriately early point. R/DM = relational/declarative memory.

We have now introduced the term as relational/declarative memory (R/DM) at its first appearance in the manuscript and have ensured that the abbreviation is used consistently thereafter.

(5) Confusing statement? "In the 40s trace-conditioned group, while the frequency of the calcium transients did not differ between WT and Cntnap2 KO mice throughout conditioning (Figure S1G), their amplitude was significantly reduced both during the presentation of the second tone and consistently across the three trace intervals in Cntnap2 KO mice (Figure 1I)."

We have revised this sentence to improve its clarity and explicitly distinguish between the frequency and amplitude of calcium transients. The revised text now makes clear that, although the frequency of calcium transients did not differ between genotypes throughout conditioning, their amplitude was significantly reduced in Cntnap2 KO mice during the second tone presentation and across the three trace intervals.

Reviewer #2 (Recommendations for the authors):

(1) The manuscript would benefit from a clearer distinction between conclusions directly supported by the data and broader interpretations. In particular, statements suggesting a shift from declarative to procedural memory systems could be presented more cautiously, as the cFos analyses provide indirect rather than causal evidence for such reorganization.

We thank the reviewer for this constructive comment. We have revised the manuscript throughout to more clearly distinguish conclusions that are directly supported by our data from broader interpretations. In particular, statements referring to a shift from declarative to procedural memory systems have been moderated.

(2) Additional clarification of the behavioral interpretation of the radial maze task would be valuable for readers who are less familiar with this paradigm, particularly regarding its relationship to relational/declarative memory and its distinction from procedural learning strategies. In addition, the results' interpretation in this paradigm is unclear to me. What is the significance of the 20-second delay in this task (why not use 10 sec or 60 sec)? I believe the paradigm tests spatial rather than temporal distant items (in contrast with trace fear conditioning that refers to temporally distant events)? Please explain further the link between the two behavioral paradigms.

We thank the reviewer for these comments.

Regarding the relationship between relational/declarative and procedural learning: The R/DM task was designed by our group (Marighetto and colleagues; Mingaud et al., 2007; Sellami et al., 2017, 2018) specifically to dissociate two learning/memory systems that can support the same behavioral output (correct arm choice) but that differ fundamentally in their underlying representations and, critically, in their flexibility.

During acquisition, an animal can solve each of the three arm-pair discriminations either by forming a flexible, relational representation of the whole spatial configuration (i.e. an allocentric/hippocampus-dependent "cognitive map" strategy, in which the reward's location is encoded relative to distal cues and to the other pairs) or by learning a set of rigid, response-based rules (i.e. an egocentric/striatum-dependent "turn left/turn right" procedural strategy tied to each specific pair).

Both strategies can produce equivalent accuracy during initial acquisition, which is why acquisition performance alone cannot distinguish them. The flexibility probe (recombining pairs A and B into a novel pair AB, without moving the reward) is the critical dissociation: only an animal that encoded the reward's location relationally, within a broader spatial map, can generalize correctly to this untrained configuration; an animal that relied on a rigid stimulus– response rule for each pair individually would be unable to solve the recombined pair above chance. Performance on pair AB is therefore the read-out of hippocampus-dependent relational/declarative memory, while the degree of left–right lateralization during acquisition (quantified by a lateralization index) is a converging behavioral signature of reliance on the egocentric/procedural system.

We have clarified this distinction in the Methods and Results to facilitate interpretation of the paradigm.

Regarding the 20-s delay and its relationship to trace fear conditioning: The reviewer is correct that the R/DM task and TFC differ in the information being integrated: the R/DM task involves spatially distinct discrimination episodes, whereas TFC involves temporally separated stimuli presented in the same context. However, both tasks require information separated by an interval to be integrated into a unified memory representation, a process referred to here as “temporal binding” (Sellami et al., 2017). The 20-s inter-trial interval in the R/DM task was based on our previous work, in which this interval was shown to be sensitive to age-related deficits in temporal binding and relational memory (Sellami et al., 2017). Similarly, our TFC experiments established a longer temporal-binding capacity in young mice (up to 40 s) compared with aged mice (20 s). Thus, although the two paradigms involve different types of information, they share the requirement to maintain and integrate information across a temporal gap. The different intervals used in the two paradigms reflect their distinct task structures and demands.

(3) Although the authors address potential locomotor confounds, a brief discussion of how hyperactivity and impulsivity might influence behavioral performance in the different paradigms would strengthen the interpretation of the results.

We thank the reviewer for raising this point. We agree that distinguishing hyperactivity from impulsivity strengthens the interpretation of our behavioral phenotypes, and we have added a discussion paragraph making this distinction explicit. Briefly: in the TFC paradigm, our existing open-field data (Figure S1D) show that while Cntnap2 KO mice travel more distance and move faster, their resting time is unaffected relative to WT; so, the reduced freezing/temporal binding phenotype at the 40 s trace cannot be attributed to a general locomotor confound, since freezing (immobility) was actually comparable between genotypes during acquisition. In the R/DM radial maze task, however, we agree the phenotype is better read as impulsivity than hyperactivity per se: KO mice show shorter decision latencies specifically at the choice point (Figure 2G), with no corresponding genotype effect on their post-choice running speed for correct trials (Figure 2H), indicating that the effect is on deliberation before response rather than on general motor output. We have revised the Discussion to make this distinction, and its implications for interpreting the egocentric/procedural bias, explicit.

(4) Minor presentation issues:

- Ensure that all statistical tests, sample sizes, and post hoc comparisons are reported consistently throughout the manuscript and figure legends.

We have carefully reviewed the statistical reporting throughout the manuscript and figure legends to ensure consistency. We have clarified sample sizes, statistical tests, and post hoc comparisons where required and in the methods.

- Carefully proofread the manuscript for minor typographical and grammatical errors. For instance, a) several figures numbers indicated in the main text do not correspond to the figures the authors refers to (page 5 Figure 2I, page 6 Figure 2J and 2K, page 7 "data in fig S2"... and so on); b) rephrase (not clear) page 9: "Differences in the relative contribution of individual substructures between trained WT and Cntnap2 KO mice involved dCA1, dCA2, and dCA3, and favored the DMS, IL, and ACC."

We thank the reviewer for highlighting these issues. We have carefully proofread the manuscript and corrected all figure reference errors and typographical inconsistencies throughout the text. We have also revised the unclear sentence on page 9 to improve clarity and accuracy.

- Page 9, an additional sentence is needed to explain the significance of the bibliographical reference: "Atypicalities in declarative memory have been reported in ASD, with specific impairments in relating items (Minor et al., 2023)."

We have revised this section to clarify the significance of the cited study and its relevance to our findings. Specifically, we have expanded the sentence to highlight that impairments in relational processing in ASD may reflect difficulties in integrating individual experiences into coherent memory representations, a process that relies on hippocampal function.

Reviewer #3 (Recommendations for the authors):

Major Points:

As stated in the public review, the authors need to thoroughly check their manuscript before submission. There are too many typos in the current version that affect the readability. For example, but not limited to:

(1) Page 5, "Figure 2I" should be "Figure 1I"; "Figure 2J" should be "Figure 1J"; "Figure 2K" should be "Figure 1K".

(2) The figure legends of Figure 1J-L are missing.

(3) Page 8, "In contrast, Cntnap2 KO mice showed no learning-induced dCA1 activation together with hypoactivation of dCA3 in both naïve and trained groups compared to WT mice (Figure S2A)", should it be dCA2 instead of dCA1? Please double-check the manuscript.

We thank the reviewer for highlighting these issues. We have thoroughly checked the manuscript and corrected all typos, figure reference errors, and fixed figure legend information. Regarding the point (3), we did mean to describe the lack of training-induced activation of dCA1, but failed to reference back to Figure 3, likely causing the confusion. This has now been clarified. The manuscript has also been carefully proofread to improve clarity and readability.

Minor points:

There are two duplicate rows (two wt) in their raw datasheet, Fig3BCD & S2 and Fig3EF. Please correct them in case of further problems.

WT Naive 178.736 103.821 250.703 185.229 437.575 50.535 266.254 306.187 306.187 WT Naive 178.736 103.821 250.703 185.229 437.575 50.535 266.254 306.187 306.187

WT Naive 46.87 37.94 15.19 8.57 4.98 12.02 8.88 20.98 2.42 12.77 14.68 14.68 WT Naive 46.87 37.94 15.19 8.57 4.98 12.02 8.88 20.98 2.42 12.77 14.68 14.68

We thank the reviewer for noticing this mistake. These duplicated rows have been corrected.

Version 2

Abstract

Growing evidence implicates the hippocampus in the pathophysiology of autism spectrum disorder, particularly in the domains of social interactions and cognition. Yet, the mechanisms driving the hippocampal-dependent cognitive atypicalities in autism remain poorly defined. Here, we characterized how dysfunction of the CA1 subfield of the dorsal hippocampus drives abnormalities in critical components of declarative memory. Using trace fear conditioning in the Cntnap2 knockout mouse model of autism, we found that capabilities to retain the association of temporally distant stimuli (i.e. temporal binding) were reduced relative to wildtype mice. Fiber photometry and optogenetic experiments demonstrated that reduced CA1 activity during temporal gaps underlies this impairment, linking CA1 dysfunction to a deficit in long-term memory retention in autism. Using a relational/declarative memory task, we also revealed a deficit in flexible spatial memory, and a preferential use of egocentric learning strategy. This inflexible learning strategy resulted from an imbalance between memory system activity, promoting frontostriatal-dependent procedural learning instead of dorsal CA1/hippocampus-dependent relational/declarative memory. Overall, this study establishes dorsal CA1 dysfunction as a circuit-level mechanism underlying cognitive inflexibility in autism, providing a neurobiological framework for hippocampal-dependent memory deficits in the condition.

Introduction

Autism spectrum disorder (ASD) is a lifelong neurodevelopmental disorder comprising difficulties in social behavior, atypical communication skills, and the presence of restricted or repetitive behaviors or interests (DSM-5 and Association American Psychiatric, 2013). Beyond these core features, individuals with ASD present with difficulties in cognitive domains that can impact daily functioning and adaptability. In particular, people with ASD present with atypicalities in episodic/declarative memory (Bal et al., 2015), the ability to recall specific events and experiences (i.e., what happened, when, and where). However, the underlying neurological mechanisms of declarative memory impairments in ASD have yet to be fully elucidated.

While most research has focused on the dysregulation of the prefrontal cortex and the basal ganglia in ASD, recent studies have highlighted that hippocampal defects are key to the social domain of this phenotype (Banker et al., 2021). In addition, the hippocampus is known to play a critical role in declarative memory by forming temporally and spatially organized relational representations of past experiences, allowing flexible expression of memories (Eichenbaum, 2004). The CA1 subfield plays a key role in this relational process by sustaining the ability to link temporally distant events, known as temporal binding (Eichenbaum, 2014; Sellami et al., 2017). In autistic individuals, difficulties in the binding of episodic memories have been proposed in autistic individuals while memory for items alone remains relatively spared, positing a selective weakness in hippocampus-dependent linking of items and contexts (Bowler et al., 2011; Minor et al., 2023). In parallel with this relational hypothesis of memory defects, time estimation tasks in individuals with ASD have shown that disruptions in integrating temporal information affect the development of temporal relationships (Buonomano et al., 2023; Maister and Plaisted-Grant, 2011; Vogel et al., 2022). In this context, we hypothesized that CA1 activity underlying temporal processing and temporal linking is impaired in ASD, leading to declarative memory defects.

Here, we examined this hypothesis in the Cntnap2 knockout (KO) mouse model of ASD, which exhibits autism-like phenotypes, including social behavior deficits (Peñagarikano et al., 2011; Peñagarikano and Geschwind, 2012), hippocampal alterations, and CA1 circuit-level changes and impairments in spatial discrimination (Paterno et al., 2021). Animals were assessed in two behavioral paradigms: trace fear conditioning (Pavlov, 1927; Sellami et al., 2017), a simple temporal binding-dependent memory task allowing the simultaneous recording and manipulation of CA1 activity; and a radial maze task modeling the complexity of temporal binding-dependent relational/declarative memory (R/DM). We demonstrate that Cntnap2 KO mice present with mistuned dCA1 activation and abnormal reorganization of memory systems during temporal binding-dependent memory formation.

Results

Cntnap2 KO mice demonstrate reduced temporal binding capability

We tested the temporal binding capabilities of the Cntnap2 KO mouse model using the trace fear conditioning paradigm (TFC; Figure 1A), in which mice are required to memorize the association between a tone (Conditioned Stimulus) and a mild electric foot-shock (Unconditioned Stimulus), separated by a temporal gap. Successful conditioning across this gap is thought to depend on the organism maintaining a decaying memory representation, or trace, of the CS after its offset; accordingly, this gap is conventionally termed the trace interval, and the paradigm trace conditioning (Pavlov, 1927; Sellami et al., 2017). In the conditioning stage, independent groups of mice were exposed to three tone-shock pairings with a trace interval between 5 and 40 s (i.e. T5, T20, or T40; Figure 1A). All groups showed intact conditioning abilities, demonstrated by high and increasing freezing times (Blanchard et al., 1986) during the tones and traces periods (Figure 1B, Figure S1A). Twenty-four hours later, the animals were re-exposed to the conditioned tone in a neutral context (i.e. tone test) to assess 24h retention capacities of the tone-shock association. Retention was reflected in fear responses to the tone and trace compared with before and after the tone (i.e., conditioned fear response in Figure 1C, % time freezing in Figure S1B). Elevated freezing during the pre-tone period at T40 may reflect fear generalization (Figure S1B, T20 vs. T40 (Sellami et al., 2017)). Interestingly, Cntnap2 KO mice also displayed a strong and specific freezing response to the tone for up to a 20 s trace interval (Figure 1C, T20), but not for 40 s (Figure 1C, T40; Figure S1B, T40). The animals were then exposed to the conditioning context alone to evaluate contextual memory. As previously demonstrated (Al Abed et al., 2020b; Sellami et al., 2017), WT mice displayed efficient retention of the tone-shock association up to a 40-s trace interval, illustrated by a strong and specific conditioned fear response to the tone in a neutral context (Figure 1D). These results were replicated in both control and Cntnap2 KO littermates (Figure S1C). Altogether, these results indicate that Cntnap2 KO mice exhibit reduced temporal binding capacity compared to WT mice, which compromises the 24h retention of the tone-shock association.

Reduced retention of temporal associations is associated with altered calcium responses during trace fear conditioning acquisition in Cntnap2 KO mice

A. Experimental design: Trace Fear Conditioning includes a conditioning session in which mice have to associate a tone and a shock, separated by a time interval (trace). Independent groups of mice were trained with trace intervals of 5, 20, or 40 seconds. B. Freezing levels (percent) during the third tone of the conditioning session for WT (black) and Cntnap2 KO (purple) groups, conditioned with distinct trace durations (T5: WT n=10; Cntnap2 KO n=11; T20: WT n=12; Cntnap2 KO n=11; T40: WT n=12; Cntnap2 KO n=14). C. Conditioned fear response during the tone test (i.e., ratio of freezing before the tone compared to during the tone and the trace, see methods), 24h post-conditioning with distinct trace durations (Genotype X Trace interaction: F2,65=11.278; p<0.0001; T40: Genotype: F2, 24=6.944; p=0.0163). D. Freezing levels in the conditioning context (context test) for WT (black) and Cntnap2 KO (purple) groups, 24h post-conditioning. E. Fiber photometry. Left: scheme depicting the fiber optic implantation targeting the dorsal CA1 subfield of the hippocampus (dCA1). Right: Representative image of the optic fiber placement, in the pyramidal cell layer of the dCA1, scale: 100µm. Calcium activity was recorded using the calcium indicator GCaMP6f (green). F. Freezing levels (percent) during the third tone of the conditioning session for WT (black) and Cntnap2 KO (purple) groups, conditioned with distinct trace durations (T20: WT n=8; Cntnap2 KO n=8; T40: WT n=11; Cntnap2 KO n=10). G. Conditioned fear response during the tone test (i.e., ratio of freezing before the tone compared to during the tone and the trace, see methods), 24h post-conditioning with distinct trace durations (T40: Genotype: F1, 19=50.143; p<0.0001). H. Left: Peak amplitude of Ca2+ transients during the 3 tone presentations of the conditioning under a 20s trace-interval (ns). Right: Peak amplitude of Ca2+ transients during the 3 trace intervals of the conditioning under a 20s trace-interval (Genotype x Trace interaction: p=0.0476; WT n=8; KO n=8, trace 1: ns (p=0.0675)). I. Left: Peak amplitude of Ca2+ transients during the 3 tone presentations of the T40 conditioning session (Genotype x Tone interaction: p=0.0247; Tone: p=0.0053; WT n=11; KO n=10). Right: Peak amplitude of Ca2+ transients during the 3 trace intervals of the T40 conditioning session (Genotype: p=0.0408; WT n=11; KO n=10). J. Optogenetics: Left panel: scheme depicting the fiber optic implantations bilaterally targeting the dorsal CA1 subfield of the hippocampus (dCA1). Right panel: Representative image of the optic fiber placement, in the pyramidal cell layer of the dCA1, scale: 100µm. Pyramidal neurons activity was manipulated only during the trace interval using Channel Rhodopsin (ChR2). K. Freezing levels (percent) during the third tone of the conditioning session for GFP-injected WT (black, n=6), GFP-injected Cntnap2 KO (purple, n=7), and ChR2-injected Cntnap 2 KO (light purple/white, n=11) groups, conditioned with 40s trace duration. L. Conditioned fear response during the tone test (i.e., ratio of freezing before the tone compared to during the tone and the trace, see methods), 24h post-conditioning with distinct trace durations (Groups: F2, 21=15.576; p<0.0001; WT-GFP vs KO-GFP: p=0.0010; KO-GFP vs KO-ChR2: p<0.0001). Data shown at mean ± SEM. *: p<0.05; **: p<0.01; ***: p<0.001; * represents genotype effect. # represents genotype x variable interaction.

As freezing is the absence of movement, the well-reported hyperlocomotion in Cntnap2 KO mice (Peñagarikano et al., 2011); Figure S1D) could represent a potential confounding factor. However, we observed comparable levels of resting time (Figure S1D, right panel) and similar freezing levels between WT and Cntnap2 KO mice during acquisition, regardless of the trace interval (Figure S1A), as well as during memory tests under control conditions (Figure S1B; T5, T20, and T40). These findings indicate that locomotor activity alone cannot account for the observed differences in freezing. Taken together, our results demonstrate that Cntnap2 KO mice exhibit reduced temporal binding capacity during conditioning, which does not affect acquisition but impairs the 24h retention of the tone-shock association.

Mistuning of dorsal CA1 activity during temporal binding in Cntnap2 KO mice

Whilst alterations in item binding in ASD have already been reported, the underlying mechanisms have not been described (Minor et al., 2023). As the sustained activity of dCA1 neurons during time intervals between stimuli is necessary for long-term retention of their association(Al Abed et al., 2020b; Sellami et al., 2017), we examined dCA1 activity during acquisition in Cntnap2 KO mice. We recorded calcium transients during encoding and recall of trace fear conditioning, using fiber photometry in animals trained under a 20s or 40s-trace interval (Figure 1E). Both WT and Cntnap2 KO mice implanted with an optic fiber could acquire the tone-shock association, reaching high levels of freezing after three tone-shock pairings (Figure 1F, Figure S1E). During conditioning, in both T20 and T40 conditions, regardless of genotype, the frequency of the calcium transients in response to the tone significantly decreased across tone presentations, suggesting neural adaptation to successive tone exposure (Figure S1F, G). Cntnap2 KO mice trained under the 20s trace interval displayed no difference compared to WT in the frequency of calcium transients during the tone and the trace interval (Figure S1F), while the amplitude tended to be higher during the first trace interval in Cntnap2 KO mice (Figure 1H). In the 40s trace-conditioned group, the frequency of the calcium transients did not differ between WT and Cntnap2 KO mice throughout conditioning (Figure S1G). In contrast, the calcium transient amplitude was significantly reduced both during the presentation of the second tone (Figure 1I, left) as well as throughout the three trace intervals in Cntnap2 KO mice (Figure 1I, right). Overall, these results suggest a lack of synchronized neuronal activity during encoding, as previously reported (Simon and Wallace, 2016). During the tone test, as in Figure 1C, Cntnap2 KO mice displayed appropriate retention of the tone-shock association when conditioned under a 20s-trace interval while the ones trained under a 40s-trace interval failed to express a fear response to the tone (Figure S1H, I). This indicates impaired retention of trace fear memory and temporal binding ability, relative to WT mice. The findings also suggest that sustained activity of dCA1 neurons during encoding of the three trace intervals, which is higher in WT than in Cntnap2 KO mice, supports temporal binding and enables adequate retention of the association between the 40s-distant tone and shock despite their temporal separation without affecting initial acquisition. These results further support the idea that the reduction in temporal binding capacity in Cntnap2 KO mice stems from a lack of sustained dCA1 activity during encoding and recall of the tone-shock association.

In WT mice, we have previously demonstrated that the activity of dCA1 principal neurons during the trace interval (but not outside of the trace interval) is required for long-term retention, but not for encoding of temporal associations (Sellami et al., 2017). To further confirm a causal link between sustained dCA1 activity during temporal gaps and 24h retention in Cntnap2 KO, we optogenetically activated principal neurons specifically during the 40s trace interval of the conditioning session using ChannelRhodopsin (ChR2; Figure 1J), while leaving neuronal activity unaltered during the remainder of acquisition and the 24h test. During the conditioning session, all groups displayed increased freezing across the three tone-shock pairings, indicative of efficient acquisition (Figure 1K, Figure S1J). Twenty-four hours later, ChR2-expressing Cntnap2 KO mice demonstrated a specific freezing response to the tone and trace, unlike GFP-expressing Cntnap2 KO mice (Figure 1L, Figure S1K). These results demonstrate that dCA1 neuronal activity during trace intervals drives temporal binding in Cntnap2 KO mice.

Egocentric learning and inflexible long-term memory in Cntnap2 KO mice

Following the observation that Cntnap2 KO mice present reduced dCA1 activity associated with temporal binding capabilities, we examined potential atypicalities in declarative memory. We used a radial maze task that models the complexity of temporal binding-dependent relational/declarative memory (R/DM) (Sellami et al., 2018). This task assesses the characteristic flexibility of R/DM expression, which relies on the dCA1-dependent ability to relate temporally distant events (Sellami et al., 2017).

To do so, the R/DM task is divided into two stages: an acquisition phase and a flexibility test (Figure 2A). In the acquisition stage, mice are trained daily to learn the food-reward location within 3 separate pairs of adjacent arms. Critically, this task structure requires mice to encode each pair as a distinct, temporally separated experience, rather than allowing continuous comparison between pairs. Cumulative evidence shows that declarative flexibility relies on a hippocampus-dependent relational representation of separate experiences formed for each pair during the acquisition phase (Al Abed et al., 2020b; Mingaud et al., 2007; Sellami et al., 2017). After reaching the learning criterion (>75% correct choices of the rewarded arm), declarative memory is then probed by testing whether animals can recombine two of these separately learned pairs (A and B) into a novel configuration (pair AB) without any change to the reward location (Figure 2A). Successful performance requires that the two pairs were encoded relationally, within a shared spatial representation, allowing this new relation to be inferred; poor performance, in contrast, suggests animals relied on rigid, pair-specific response rules that do not generalize to novel arm combinations. To impose a temporal binding demand on relational memory acquisition, the presentation of each pair is separated by a 20s inter-trial interval. This timing was empirically determined in (Sellami et al., 2018, 2017).

Cntnap2 KO mice display impaired relational/declarative memory.

A. Experimental design: Relational/Declarative memory task. ITI: Inter-trial interval; +: represent reward location within each pair of arms. B. Average correct answers for WT (black; n = 15) and Cntnap2 KO mice (purple; n = 14) for pairs A, B, and C during the 5 first days of acquisition (left panel; day (D); D4: Genotype: p=0.0149), and the last 5 days of acquisition (right panel; D-5: 5 days before they reach learning criterion; D-1: Genotype: p=0.0082). C. Comparison of the correct answers during the last day of training (D-1) averaged on pairs A and B, and during the flexibility test on the recombined pair AB (Genotype X Performance interaction: F1,27=12.639, p=0.0014; Test, Genotype: p=0.0239). Inset: D-1 vs. Test performance index on pairs AB (Genotype: p=0.0022). D. Comparison of the correct answers on pair C during the last day of training (D-1) and the flexibility test. Inset: D-1 vs. Test performance index for Pair C. E. Average performance on the new pair N during the flexibility test. F. Maximum lateralization index during stage 1 (absolute value of the average performance between pairs A and B (left baited) minus Pair C (right baited); Genotype: p=0.0153). G. Average decision latencies for correct (+) and incorrect (-) answers throughout training (Genotype X Response interaction: F1,31=7.960, p=0.0083). H. Average runtimes for correct (+) and incorrect (-) answers throughout training (Resp -, Genotype: p=0.0500). Dashed line illustrates chance level. Data shown at mean ± SEM. *: p<0.05; **: p<0.01; # shows genotype x performance interaction.

Both wild-type (WT) and Cntnap2 KO mice were able to acquire the task (Figure 2B). Cntnap2 KO mice tended to require more days to reach criterion (average number of days: WT: 8.06, KO: 10.86, ns, raw data in Figure S3) and reached higher performance levels at the end of training (Figure 2B, right panel).

In the flexibility test, their performance dropped significantly between the last training session and the flexibility test (stage 2 of the R/DM task), and Cntnap2 KO mice performed significantly worse than WT (Figure 2C - Recombined pair AB). In contrast, both groups performed similarly on the control pairs: the unchanged pair C, with high performance (Figure 2D), and the novel pair N, at chance levels (Figure 2E).

What could explain this lack of flexibility? We have previously shown that the flexibility component of R/DM depends on the learning strategy used during the acquisition stage: either allocentric spatial mapping (relational) or egocentric (procedural), the latter leading to lateralization. Indeed, when the use of spatial cues is impaired, as in cases of hippocampal damage or dysfunction, mice adopt a left-right egocentric strategy to memorize reward locations leading to inflexible memory (Mingaud et al., 2007; Sellami et al., 2017). Therefore, the chance-level flexibility performance of Cntnap2 KO mice suggests a bias towards a procedural learning strategy, in line with the well-documented prominence of behavioral inflexibility (Uddin et al., 2015) and the use of egocentric learning strategies in ASD (Brunner et al., 2015; Ring et al., 2018). This was confirmed by analyzing lateralization during acquisition by comparing performance on left-baited pairs (pairs A and B), vs. right-baited pair C (Methods, Figure 2F). Cntnap2 KO mice displayed significantly more lateralization than WT, meaning they used an egocentric/procedural strategy to learn the position of the food reward rather than using the spatial mapping strategy observed in WT mice. We found no significant correlation between the use of egocentric strategy and the number of days to reach criterion (Data in Figure S3). Notably, performance during Day 4 of acquisition in WT mice was significantly negatively correlated with their use of lateralized strategy (r2 = -0.518, p=0.0471), which was not seen in Cntnap2 KO mice. Together, this further confirms that Cntnap2 KO mice employ a lateralized strategy to learn the task, which impacts both task performance (Figure 2B, day 4), and memory flexibility (Figure 2C).

Interestingly, Cntnap2 KO mice had shorter decision times than WT (Figure 2G), consistent with the well-demonstrated impulsivity observed in ASD (DSM-5 and Association American Psychiatric, 2013). In addition, WT mice demonstrated a longer time to enter a non-baited arm than a baited arm, suggesting that they evaluate reward location before entering the arm, likely relying on deliberation and precise spatial representation. In contrast, Cntnap2 KO mice showed similarly short latencies regardless of their choice, suggesting impulsive, automatic responses and impaired spatial representation of reward information. In addition, once a choice was made, their running time (from the central platform to the end of the arm) was longer than that of WT mice (Figure 2H), implying that while the deliberative phase is shortened or absent in KO mice, there may be a compensatory adjustment following impulsive decisions.

Altogether, these findings indicate that Cntnap2 KO mice exhibit impaired flexible memory expression, consistent with a relational/declarative memory deficit and increased reliance on an egocentric/procedural learning strategy.

Potential shift towards striatum-centered activities in Cntnap2 KO mice

Each learning strategy is known to depend on partially distinct brain systems: hippocampus-dependent systems supporting allocentric spatial/relational learning, and striatum-based systems leading to egocentric/procedural learning (White and McDonald, 2002). To assess learning-induced activity in these different structures, we quantified the levels of the cellular activity marker cFos after day 3 of R/DM task acquisition in structures involved in memory (i.e., the dorsal hippocampus [dCA1, dCA2, dCA3, the dorsal dentate gyrus (dDG)]; the dorso-medial (DMS) and the dorso-lateral striatum (DLS); the prefrontal cortex [infralimbic (IL), prelimbic (PrL), and anterior cingular (ACC)]; Figure 3A, Figure S2).

Learning-induced activity and functional connectivity in Cntnap2 KO mice.

A. Experimental design: cFos levels were quantified 90min after the 3rd training session of the Relational/Declarative memory task. ITI: Inter-trial interval; + represents reward location within each pair of arms (WT Naïve: n=5; WT Trained: n=7; Cntnap2 KO: n=5; Cntnap2 KO Trained: n=7). B. number of cFos-positive neurons (cell density/mm2) in the dorsal dCA1 (WT naïve vs Trained: p=0.0491; Trained WT vs. KO: p=0.0129) with representative images of the cFos staining (Magenta) in dCA1, scale: 100µm, for naïve (light colors) and trained (dark colors) WT (grey/black) and Cntnap2 KO (purple) groups. C. number of cFos-positive neurons (cell density/mm2) in the Dorso-Medial Striatum (DMS; Naive WT vs. KO: p=0.0189; KO Naïve vs Trained: p=0.0085) with representative images of the cFos staining (Magenta) in DMS, scale: 200µm. D. Percentage of total activity: activity of each structure over to the total activity measured (percentage; Trained WT vs. KO: Hippocampus: p=0.00194; PFC: p=0.00107; Striatum: p=0.0382). E. Percentage of total activity: activity each subfield within the total network, in WT mice (naïve vs trained, left panel; dCA1: p=0.0242), in Cntnap2 KO mice (Naïve vs trained, middle panel), and in trained mice (WT vs Cntnap2 KO mice, dCA1: p=0.009; dCA2: p=0.0174; dCA3: p=0.0274; ACC: p=0.0438; DMS: p=0.0350; IL: p=0.0267; PrL: p=0.0476; right panel, same data as in left and middle panels). F. Pairwise between-structure correlation strength of cFos cell density. Data shown as mean ± SEM. *: p<0.05; **: p<0.01, ***: p<0.001.

As previously shown (Al Abed et al., 2020b; Sellami et al., 2017), we found a significant activation of dCA1 in trained WT mice compared to naïve WT mice (Figure 3B), consistent with the use of an allocentric/spatial learning strategy (White and McDonald, 2002). In contrast, Cntnap2 KO mice showed no learning-induced dCA1 activation (Figure 3B), together with hypoactivation of dCA3 in both naïve and trained groups compared to WT mice (Figure S2A). This reduced hippocampal activation may contribute to the observed deficit in temporal binding and spatial learning capabilities in this task. Moreover, Cntnap2 KO mice displayed a significant training-induced activation of the DMS compared to their naïve controls, a pattern not observed in WT mice (Figure 3C). No other significant changes were detected in the hippocampus (i.e. dCA2, DG; Figure S2A), in the DLS (Figure S2B) or in the prefrontal cortex regions (i.e.IL, PrL, ACC; Figure S2C). Together, these findings suggest increased engagement of DMS-associated networks during the egocentric learning strategy adopted by Cntnap2 KO mice.

To further dissect the balance between memory systems, we analyzed the relative contribution of each structure within the total activation of the system (Figure 3D-E). After training, we found a disproportionate representation of the PFC and striatum relative to the hippocampus in Cntnap2 KO mice compared to WT (Figure 3D). Indeed, compared to WT naïve mice, WT trained mice exhibited an increase in the weight of the dCA1 in the network activity, likely supporting temporal binding and, through ensuing relational organization of the 20s-distant pair presentations, the formation of flexible memory. In contrast, no such reorganization of network activity was observed in Cntnap2 KO mice with training (Figure 3E). We observed differences in the relative contribution of individual substructures between trained WT and Cntnap2 KO mice. Specifically, training was associated with greater involvement of the dCA1, dCA2, and dCA3 in WT mice, whereas Cntnap2 KO mice showed greater involvement of the DMS, IL, and ACC. Overall, we observed compensatory activation of the frontal-striatal structures concomitant with hippocampal hypoactivation in the Cntnap2 KO model.

In addition to the learning-induced activity of individual structures, pairwise correlation of activity between structures enhances our understanding of network-wide synchronization. While naïve WT mice exhibited mainly positive correlations between the striatum, PFC, and hippocampus, with the exception of the dDG which was negatively correlated with the PFC, naïve Cntnap2 KO mice displayed predominantly negative correlations (Figure 3F).

This indicates that, at baseline, inter-structural relationships are dysregulated in Cntnap2 KO mice, potentially limiting the recruitment of a hippocampus-based learning strategy. Altogether, the analysis of learning-induced brain activity suggests a shift towards striatum-centered activity and connectivity in Cntnap2 KO mice, likely promoting an egocentric learning strategy and resulting in inflexible memory.

Discussion

This study establishes a novel mechanistic framework that builds on prior reports of hippocampal alterations in Cntnap2 KO mice (Paterno et al., 2021). It demonstrates that compromised CA1 hippocampal activity during encoding disrupts temporal binding, which, by impairing the relational processing of temporally distant events, leads to deficits in retention of declarative memory. Moreover, hippocampal dysfunction in Cntnap2 KO mice appears to shift learning away from hippocampus-dependent strategies and toward egocentric strategies.

Interplay between hippocampus and striatum in memory deficits in Cntnap2 KO mice

Individuals with ASD show atypicalities in declarative memory, characterized by impaired relational binding, i.e., the ability to link discrete items or events into a unified representation, while memory for individual items in isolation appears comparatively preserved (Minor et al., 2023). Because many real-world relations must be formed across a temporal gap (e.g., linking an event to its outcome, or one experience to another encountered later), we propose that this relational impairment may arise, at least in part, from a deficit in temporal binding: the hippocampus-dependent process of maintaining and integrating information across time to construct coherent, flexible memories.

Here we demonstrate a reduction in dCA1-dependent temporal binding capacity in Cntnap2 KO mice, which could contribute to the declarative memory impairment by disrupting the relational organization of temporally distant events. Indeed, in the trace fear conditioning experiments Cntnap2 KO mice showed efficient temporal binding with time intervals up to 20s, compared with 40s in WT mice, and this reduction was associated with a reduction of dCA1 activity during the encoding of the 40s-trace fear conditioning. Temporal integration of stimuli relies on hippocampal activity, in particular that of dCA1 (Buzsáki and Tingley, 2018; Chowdhury et al., 2005; Sellami et al., 2017; Tsien et al., 1996). Similarly, we found an impaired activation of the dCA1 in Cntnap2 KO mice during the acquisition of the Relational/Declarative Memory (R/DM) task (Figure 3B).

The lack of training-induced dCA1 activation likely prevents fine-tuned temporal integration of distant events, leading to flexibility deficits (Chang and Gold, 2003; Etchamendy et al., 2003; Mingaud et al., 2007; Sellami et al., 2017). Moreover, analysis of broader memory systems’ activity during encoding revealed an aberrant recruitment of the DMS in the Cntnap2 KO mice, along with an altered pair-wise correlation between activation of related brain regions (Figure 3F). While WT mice exhibited predominantly positive correlations, Cntnap2 KO mice showed negative correlations among the DLS, prefrontal cortex and hippocampus. This pattern was also observed in naïve groups, suggesting that the bias in ASD towards lateralized learning strategy stems from baseline mistuning of functional connectivity between brain structures necessary for spatial learning in ASD.

This shift toward a striatum-centered system in ASD also implies a predominance of unconscious encoding, in which relations between items are acquired via stimulus–response associations, leading to inflexible, habit-like actions (Ferbinteanu, 2016). Whether the increased reliance on corticostriatal pathways reflects a compensatory adaptation to impaired hippocampal processing or a direct consequence of Cntnap2 deficiency remains an open question. Cntnap2 is broadly expressed across cortico-hippocampal and corticostriatal circuits and has been implicated in neuronal connectivity, synaptic function, and network synchronization (Lazaro et al., 2019; Scott et al., 2019; Selimbeyoglu et al., 2017). It is therefore possible that the altered engagement of corticostriatal networks observed here is not solely secondary to reduced hippocampal activity, but also reflects primary circuit alterations caused by the genetic mutation. Future studies directly examining corticostriatal activity and manipulating these circuits during learning will be required to disentangle primary genetic effects from compensatory network adaptations. However, additional mechanisms may contribute to the phenotype. Cntnap2 KO mice exhibit other neural and behavioral alterations (Peñagarikano et al., 2011), including epileptic seizures and hyperlocomotion, suggesting that CA1 hypoactivity is likely only one of several factors influencing performance. Similarly, although previous work demonstrated that dCA1 activation rescues temporal binding only when delivered during the trace interval (Sellami et al., 2017), future studies should determine whether the same temporal specificity is maintained in the Cntnap2 KO model.

Decoding spatial and temporal maps in the Cntnap2 KO mouse model of autism

The CA1 region is recruited with other hippocampal areas to form and maintain spatial, temporal and social maps, which are mental relational representations of our environment (Barrientos and Tiznado, 2016). CA1 neurons contribute to encoding and recalling spatial information, enabling individuals to navigate and understand their surroundings. Specific “time cells” create a temporal framework that supports episodic memory, encoding the passage of time and the prediction of future events (Eichenbaum, 2014). Recent research also suggests that CA1 contributes to social cognition through so-called “social maps” which process and recall social information that is essential for effective social behavior and social dynamics (Zhang et al., 2022). Yet, little is known about how cognitive deficits evolve over time and their long-term effects in ASD. While the present study focused on dCA1 activation during acquisition to assess how dCA1 binds distinct stimuli into unified representations, future investigations directly examining the causal role of dCA1 activity in memory maintenance and retrieval (e.g. ArchT-mediated silencing during recall) will be essential for a comprehensive understanding of dCA1’s role in memory processes in ASD, as well as for exploring how specific maps are formed and maintained in this condition.

The role of dCA1 in spatial representations has been profusely documented. Consistently, we recently demonstrated the central role of dCA1 in the maladaptive contextual encoding in post-traumatic stress disorder (PTSD), i.e., the degradation of declarative memory surrounding a highly stressful event (Al Abed et al., 2020a; Brewin, 2011; DSM-5 and Association American Psychiatric, 2013). Interestingly, growing evidence reports the propensity to develop PTSD in ASD (Al Abed et al., 2024; Rumball et al., 2020), suggesting that reduced dCA1 activation may impact cognition beyond everyday learning. Together, these studies further underscore the hippocampus as a promising target to alleviate ASD pathophysiology.

To obtain a deeper understanding of the hippocampus computation underlying memory formation, we analyzed the correlation between neuronal dynamics and behavioral outcomes within the fear memory test. Fiber photometry recordings in the dCA1 showed that whilst WT mice displayed sustained hippocampal activity during trace interval (Figure 1I), hippocampus-dependent memory systems were not similarly activated in the Cntnap2 KO mice (Figure 3B, E). We also found a marked decrease in the amplitude of calcium transients in Cntnap2 KO mice (Figure 1I). This profile was not observed when Cntnap2 KO mice were trained within their capacity (i.e. 20 s trace interval; Figure 1H). However, because fiber photometry provides a population-level measure of activity, it does not resolve the contribution of individual CA1 neurons or underlying cellular and synaptic mechanisms. Future studies using approaches with single-cell or subcellular resolution will be required to determine how these mechanisms contribute to the altered CA1 dynamics observed here.

Overall, these results suggest that the deficits in temporal binding and declarative memory originate from reduced hippocampal neuronal synchronization and dynamics. An additional possibility is that sustained CA1 activity during the trace interval is supported not only by recurrent network interactions but also by intracellular mechanisms within CA1 pyramidal neurons. Persistent dendritic calcium signaling and CaMKII-dependent processes, including dendritic, delayed and stochastic CaMKII activation (Jain et al., 2024) and short-term associative plasticity of calcium dynamics (Caya-Bissonnette et al., 2023; Caya-Bissonnette and Béïque, 2024), have been proposed to support prolonged information retention. Determining the relative contributions of these intracellular and circuit-level processes will require future investigations with single-cell and subcellular resolution, as the network-level approaches used in the present study did not allow us to distinguish whether the sustained activity observed here arises from these cellular mechanisms, recurrent circuit dynamics, or a combination of both. In the same line, a recent study reported that neurons allocated to the memory trace displayed freezing-specific activity compared to non-allocated neurons during contextual fear conditioning (Mocle et al., 2024). While our study design does not provide these details, the lack of active cells in the dCA1 during acquisition of the relational/declarative memory task could imply memory trace allocation to the striatum rather than the hippocampus (Rashid et al., 2016). Further experiments are needed to determine whether atypical engram allocation is linked to learning strategies in ASD.

Dissociating hyperactivity from impulsivity in Cntnap2 KO mice

As Cntnap2 KO mice display well-documented hyperlocomotion (Peñagarikano et al., 2011; Figure S1D), a general increase in motor output represents a potential confound for several of the behavioral readouts in this study. We addressed this directly for the TFC paradigm, where resting time did not differ between genotypes despite increased total distance and velocity in the open field (Figure S1D), and where freezing levels during acquisition were comparable across trace intervals and genotypes (Figure S1A). These findings indicate that the genotype-specific loss of retention at the 40 s trace interval reflects a deficit in temporal binding rather than an inability to freeze.

However, in the R/DM task our data more directly implicate impulsivity, rather than hyperactivity, as a contributing factor to task performance. Cntnap2 KO mice showed shorter decision latencies specifically at the choice point of the maze (Figure 2G), and, critically, this latency reduction was similar regardless of whether the eventual choice was correct or incorrect. In contrast, WT mice took longer to enter non-baited arms, consistent with deliberation over reward location before committing to a choice. This pattern is consistent with premature, automatic responding rather than a general increase in locomotor drive: KO mice were in fact slower once a choice had been made (i.e. longer runtimes on both incorrect and correct trials; Figure 2H), arguing against a simple speed-accuracy effect driven by hyperactivity.

This distinction is conceptually important, as impulsivity, characterized by premature responding and reduced response control, is a well-documented feature of ASD in both clinical and preclinical settings (DSM-5 and Association American Psychiatric, 2013), including in other autism-relevant mouse models assessed with operant impulse-control paradigms. Examples include the 5-choice serial reaction time task (CSRTT), where increased premature responding has been reported independently of general locomotor activity (McTighe et al., 2013). We therefore interpret the shortened decision latencies in Cntnap2 KO mice as reflecting reduced deliberative evaluation of spatial/reward information prior to responding, which may compound, but is dissociable from, the reduced hippocampal engagement and egocentric strategy bias described above.

We note, however, that our task was not designed to formally quantify impulsivity (e.g., via premature responses during an imposed delay, as in the 5-CSRTT), and future studies directly assessing response inhibition in Cntnap2 KO mice will be required to determine whether this impulsive-like phenotype is mechanistically linked to, or independent of, dCA1 hypofunction.

Aged-like performance in Cntnap2 KO mice suggests early cognitive decline

In this study, we demonstrate the disruption in hippocampal and CA1 functioning in the Cntnap2 KO mice, a hallmark of memory defects observed during aging (Etchamendy et al., 2012; Leal and Yassa, 2015). Consistent with reports suggesting that people with autism are at risk of premature cognitive aging (Wang et al., 2024), we have found that young adult Cntnap2 KO mice exhibit a memory profile similar to those of aged WT mice (Al Abed et al., 2020b; Marighetto et al., 2011; Sellami et al., 2017). Compared to young WT mice, both groups showed flexibility impairments in the relational/declarative radial maze task, and both exhibited a striatum-biased learning strategy. However, the underlying mechanisms driving these deficits remain unknown.

We previously demonstrated that dCA1 activity during the temporal gaps is necessary to link separate pairs into a relational/flexible memory (Al Abed et al., 2020b; Sellami et al., 2017) and that targeted dCA1 optogenetic activation during these intervals increases temporal binding capacity, rescuing relational/declarative memory formation. Given that CA1 neurons are less synchronized in ASD, targeted dCA1 activation could alleviate memory deficits. We propose that fine-tuning dCA1 activity may normalize memory system balance, offering a novel therapeutic approach for cognitive impairments in ASD. Further delineation of associated connectomes (Hollunder et al., 2024) could provide insights into targeted brain circuit therapies.

Autism is a highly heterogeneous condition, with individuals presenting a wide range of atypical cognitive abilities. Consequently, the extent and nature of memory deficits vary considerably across the spectrum. Research has yet to fully elucidate how these deficits manifest, particularly in relation to other cognitive domains such as attention, executive functioning, and language. Sex differences are another factor to consider (Van Wijngaarden-Cremers et al., 2014). While this study included both male and female mice, we did not observe sex-specific effects, suggesting that the behavioral and cognitive phenotypes examined in the Cntnap2 KO mouse model are largely independent of sex. This is consistent with previous studies reporting no robust sex differences (Al Abed et al., 2024).

A better understanding of these individual differences is necessary to create more personalized and effective interventions. As such, further studies will have to assess these aspects in different models to fully comprehend the common cognitive mechanisms at play in autism.

Methods and material

Mice

3 to 5-month-old naive mice were collectively housed in standard Makrolon cages, in a temperature- and humidity-controlled room under a 12h light/dark cycle (lights on at 07:00), with ad libitum access to food and water. We used both males and females to minimize the number of mice produced in this study. Cntnap2+/+ (WT) and Cntnap2-/- (KO) mice (Jackson ID: #017482 (Peñagarikano et al., 2011)) were non-littermate, though both lines were originally derived from breeders of the same colony and were used for behavior and histological analyses. To minimize potential differences between strains, we regularly inter-crossed the wildtype and Cntnap2 KO colonies (i.e., every year) to generate heterozygous mice. These heterozygotes were then bred to produce either WT or KO mice for experiments. Throughout the study, we used non-littermate mice originating from the same breeders. To ensure that observed differences between WT and Cntnap2 KO were not influenced by this breeding strategy, we also assessed trace fear conditioning capacities in littermate mice and confirmed our findings using Het × Het breeding experiments (Figure S1).

All mice were bred in the same room, and maintained under identical husbandry and rearing conditions, including cage content, bedding, diet, and light/dark cycle. All experiments took place during the light phase. We replicated the behavioral experiments in two to four different batches. Every effort was made to minimize the number of animals used and their suffering. All procedures were conducted in accordance with the European Directive for the care and use of laboratory animals (2010-63-EU) and the animals care guidelines issued by the animal experimental committee of Bordeaux University (CCEA50, agreement number A33-063-099; authorization N°21248), and from the Australian National University Animal Experimentation Ethics Committee (protocol numbers A2018/66, A2020/26, A2021/43, and A2024/362).

Trace fear conditioning task (TFC)

Apparatus

Fear conditioning behavior was performed in a Plexiglas conditioning chamber (20 x 20 x 30 cm), in a brightness of 100 lux, given access to the different visual-spatial cues in the experimental room. The floor of the chamber consisted of stainless-steel rods connected to a shock generator (Imetronic, Pessac, France).

Trace fear conditioning procedure

The box was cleaned with 70% ethanol before each trial. During conditioning, each animal received 3 pairings of a tone (70 dB, 1 kHz, 30 s) and a footshock (0.3 mA, 50 Hz, 1 s). The two stimuli were separated by a trace interval of 5, 20, or 40 s, depending on the group. Each tone-shock pairing was separated by a 1min-interval.

All mice were submitted 24 hours later to the Tone retention test during which they were re-exposed to the tone alone in a dark and modified chamber [circular opaque white chamber cleaned with a 1% acetic acid solution; 2min pre-tone, 2min tone and 2min post-tone]. A context test was also performed 2 hours later, during which mice were re-exposed to the conditioning chamber alone for 6 min.

Animals were continuously recorded for off-line scoring of freezing, by an experimenter ignorant of the experimental condition/group. Tone test was scored by 2 independent scorers. Freezing is defined as a lack of all movement except for respiratory-related movements. Acquisition of the TFC was measured by the progression of freezing during the tone across the 3 tone deliveries. 24h-retention of the trace association was measured during the tone test by the evolution of percentage of freezing time. Conditioned fear response to the tone was evaluated by calculating a normalized ratio between the tone and trace periods, versus outside these periods (i.e., Pre-Tone and Post-Tone): [(Tone+Trace)/2 - (pre-Tone+post-Tone)/2] / [(Tone+Trace)/2 + (pre-Tone+post-Tone)/2].

Radial-maze task of Relational/Declarative memory (R/DM)

Apparatus

We used an open 8-arm radial maze made of grey Plexiglas, automatized by videotracking (IMETRONIC – Pessac, France). The diameter of the central platform is 30 cm, and the arms are 55 cm long by 10 cm large. Each arm is equipped with a door at its entrance and a food-pellet delivery system at its end. The doors are individually controlled (raised up or dropped down) by the computerized system which also controls pellets availability in the food tray at the end of each arm individually, according to the task. The maze is placed in an empty room containing visual cues to enable spatial discrimination.

Behavioral Procedure

During the entire procedure, animals were submitted to one daily session. Prior to memory testing, animals were habituated to the apparatus over a period of two days during which animals were allowed to move freely in the radial maze. To complete the session, mice had to visit each arm until its end. Choice accuracy was measured as the percentage of correct responses for each pair.

Stage 1: Acquisition of 3-pair discriminations

The acquisition task consisted in learning the position of the food within the maze. Indeed, each animal was assigned three adjacent pairs of arms (pairs A, B and C). In each pair, only one arm was baited as a food reward. The experiment is designed in a way that the left arms of pairs A and B are baited while the right arm of pair C is. In each trial, mice were given access to a pair (either pairs A, B, or C). A choice was made when the subject reached the food well of an arm; this also closed the door of the non-chosen arm. The trial was finished as soon as the animal returned to the central platform. The subject was then confined to the central platform for 20s before the next discrimination trial began: this constituted the inter-pair interval (ITI). This interval was selected based on previous work showing that a 20-s ITI is sufficient to challenge temporal binding and reveal deficits in relational/declarative memory in this paradigm (Sellami et al., 2018).

Each daily session consisted of 20 consecutive trials comprising alternate presentations of pairs A, B, and C according to a pseudo-random sequence. A mouse was considered to reach criterion when its overall choice accuracy was at or above 75% over two consecutive sessions given that performance in each of the three discrimination choices were at least 62% correct. The flexibility test task began the following day when criterion performance was reached. Lateralization index was quantified by using the absolute value of the subtraction of the performance on the left-baited pairs vs. the right-baited pairs. Lat = |((Pair A + Pair B) /2) - Pair C|. We then used the maximum lateralization over 3 consecutive days.

Stage 2: Flexibility probe of DM

In the test task, the position of the food in the maze did not change, but characteristic flexibility of DM expression was assessed by changing the way of presenting the arms. Indeed, in place of the pairs A and B, a pair AB was submitted to the mouse. The AB pair consisted in the combination of the two adjacent arms of the pairs A and B. It was the critical test of flexibility. Two other pairs were used as control: the pair C that remained unchanged (“unchanged learnt control”, and pair N (=new) made of the two arms non-used in acquisition (“unlearnt control”).

Fiber photometry

Surgery

Mice were injected unilaterally 4 weeks before behavioral experiments with an Adeno-Associated Virus (AAV) GCaMP6f (AAV1-hsyn-jGCaMP6f-WPRE, Addgene). We used glass pipettes (tip diameter 25-35 µm) connected to a picospritzer (Parker Hannifin Corporation) into the right side of the dorsal CA1 of the hippocampus (0.1 µl/site; AP -1.8 mm; L +1.35 mm; DV -1.4 mm). Mice were then implanted with unilateral optic fiber implant (diameter: 200µm; numerical aperture: 0.39; flat tip; Neurophotometrics). Implants were fixed to the skull with Super-Bond dental cement (Sun Medical, Shiga, Japan). Mice were perfused after experiments to confirm correct placements of fibers. Viral injections targeted the pyramidal cell layer of the dCA1.

Recordings

Data were recorded throughout the entire fear conditioning paradigmusing a Neurophotometrics FP3002 system (Neurophotometrics, San Diego, CA). Briefly, 470 nm and 415 nm LED light was bandpass filtered, reflected by a dichroic mirror, and focused onto a multi-branch patch cord (Neurophotometrics, San Diego, CA) by a x20 objective lens. Alternating pulses of 470 and 415 nm light (∼50 µW) were delivered at 60 Hz, and photometry signals were time-locked to behavior using a Bonsai workflow. Signals were analyzed using the Synaptech suite cloud service (MetaCell) and custom-made Python scripts (Code available at https://github.com/elisecr/fiberphotometry_2024). Through the Synaptech platform, 470 and 415 signals were deinterleaved. For all experiments, the exponential delay period (the first 10 seconds(Hon et al., 2025)) of the recording was removed. To correct for photobleaching, isosbestic (415nm) signals were fit to a biexponential curve to model the baseline signal with photobleaching. The fitted biexponential curve was then scaled to match the 470nm signal using non-negative robust linear regress and subtracted from the 470 trace to correct for baseline drift. Traces were then Z-scored.

Experimental traces

Z scored data were analyzed by custom python scripts to extract calcium fluorescence traces for each experiment. For the trace fear conditioning experiments, these data were aligned to files generated by the poly fear software (Imetronic, France) “.dat” files to align the beginning of the tone and shock events to the closest millisecond in the Z scored data. The timing of the tones and trace periods were extracted as described above to accurately align the data.

Peak finding

The threshold for peak analysis definition was set at 1.0. Peak analysis was conducted using a custom-built Python script to calculate peak frequency and amplitude. Code available at https://github.com/elisecr/fiberphotometry-analysis

Optogenetic Manipulation of dCA1

Surgery

Mice were injected bilaterally 5 weeks prior to behavioral experiments with an Adeno-Associated Virus (AAV) containing the excitatory opsin channelrhodopsin targeting the pyramidal cells of dCA1 (pAAV-CaMKIIa-hChR2(H134R)-mCherry (#26975), Addgene). We used glass pipettes (tip diameter 25-35 µm) connected to a picospritzer (Parker Hannifin Corporation) into the dorsal CA1 of the hippocampus (0.2 µl/site; AP -1.8 mm; L +1.3 mm; DV -1.4 mm and AP -2.5 mm; L +2 mm; DV -1.4 mm). Mice were then implanted with bilateral optic fiber implant (AP -1.8 mm; L +1.3 mm; DV -1.3 mm; diameter: 200µm; numerical aperture: 0.39; flat tip; Neurophotometrics). Implants were fixed to the skull with Super-Bond dental cement (Sun Medical, Shiga, Japan). Mice were perfused after experiments to confirm correct placements of fibers. Viral injections targeted the pyramidal cell layer of the dCA1. Optogenetic manipulation during trace fear conditioning: Mice were conditioned under a 40s trace interval condition, while dCA1 pyramidal neurons were activated during the 40 seconds of trace interval only. The light (≈3 mW per implanted fiber) was bilaterally conducted from the laser (Neurophotometrics, 450nm; 5 Hz; 5 ms laser on, 195 ms laser off) to the mice via two fiber-optic patch cords (diameter, 200 μm; Doric Lenses). The tone test was performed 24h later, mice were tethered to the patch chord, without optogenetic manipulation.

Open field

Mice were placed in a 38.5cm diameter circular arena and allowed free exploration for 5 minutes. These movements were recorded, and the mean velocity, distance travelled, and resting time were tracked and quantified using a custom MATLAB (Mathworks) code as in (Parkinson et al., 2024).

Immunohistochemistry

Naïve (homecage) and trained (90 min after the flexibility test on Day 3 for c-Fos analysis) animals were perfused transcardially with 0.01M phosphate buffered saline (PBS) to eliminate blood and extraneous material, followed by 4% paraformaldehyde (PFA). Brains were postfixed for 36h in PFA. Tissues were sectioned at 40 µm using a Leica 1000S vibratome and kept in a cryoprotective ethylene glycol solution at -20°C until processed for immunofluorescence. Sections were first washed and permeabilized in PBS-Triton 0.25%, then non-specific binding sites were blocked by immersing the tissue in 10% normal donkey serum, 2% bovine serum albumin in PBS-Triton 0.25% during 2h. Tissues were then stained using the primary antibodies overnight: mouse anti-c-Fos (1:1000; Santa Cruz). After 3x 15 min washes, we added anti-rabbit Alexa 488 (1:200; Life Technologies) secondary antibodies for 2h. After 3x 15 min washes slices were stained during 5 min with DAPI (5µM; Sigma), mounted on Livingstone slides then covered with Mowiol (Sigma) and coverslip (Thermofisher). c-Fos, staining from 2-3 slices par animals were imaged using a Nikon Axioscan confocal fluorescent microscope (20x objective). Stained sections of WT and mutant mice were imaged during the same imaging session. Immunofluorescence signals were quantified using the quPath software with routine particle analysis procedures (size= 30-500; circularity=0.30-1.00), to obtain cellular masks, divided by the area to obtain cell density per mm2.

Statistics

Data are presented as mean ± SEM. Statistical analyses were performed using GraphPad prism or StatView software for 2-way ANOVA, followed by Bonferroni post-hoc test to assess 2 variables or more (e.g., interaction between genotype/group and performance/freezing/number of cFos+ neurons) or Student’s t test, when assessing one variable only. Normality of the data was confirmed using the Kolmogorov–Smirnov test. Statistical significance was considered at p<0.05.